---
title: 'The Anopheles mosquito brain at spatial and single-cell resolution: \n Analyis of Spatial Transcriptomics Data'
author: "Johannes W. Bagnoli"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    number_sections: false
    keep_md: false
    self_contained: yes
---

# Step 1: load required packages & functions:
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = F, error = F)
```

```{r seed}
set.seed(1234)
projDir<-here::here()
```

```{r packages, results='hide'}
libs <- c("Seurat","dplyr","reshape2","stringr","spacexr","tidyverse","gatepoints","artpack", "Rfast", "colorRamp2","ggalluvial", "cowplot", "scater", "ggrepel")
lapply(libs, library, character.only = TRUE)

```

```{r functions_Deconvolution}
## Functions related to RCTD Deconvolution
### UMI cutoff for assignment
UMI_cutoff <- function(nUMI) {
  return (pmax(0.25, 2 - log(nUMI,2) / 5))
}

### Assignemnt of cell type and NR of celltypes per spot from RCTD weights
#### object: Seurat RCTD object from spacexr
ctype_assign_RCTD<- function(obect){
  norm_weights = normalize_weights(obect@results$weights)
  occur<- numeric(length(obect@cell_type_info$info[[2]]))
  names(occur) = obect@cell_type_info$info[[2]]

  cond_res<-list()
  for (j in 1:length(obect@cell_type_info$info[[2]])) {
    cell_type = obect@cell_type_info$info[[2]][j]
    res_tmp = norm_weights[,cell_type] > UMI_cutoff(obect@spatialRNA@nUMI)
    res_tmp<- res_tmp[res_tmp == TRUE]
    if (length(res_tmp)>0){
      cond_res[[j]] = data.frame(XC=names(res_tmp), celltype=cell_type)
    }  else{
      cond_res[[j]] = data.frame()
    }
  }
  
  cond_res_df<- dplyr::bind_rows(cond_res)
  
  XC_occ<- as.data.frame(table(cond_res_df$XC))
  colnames(XC_occ)<-c("XC", "Nr_Celltypes")
  
  cond_res_df_collapse<- cond_res_df %>%   group_by(XC) %>%   summarise(cell_types = paste(unique(celltype), collapse = "_"))
  final_results<-dplyr::left_join(XC_occ, cond_res_df_collapse)
  rownames(final_results)<-final_results$XC
  add_data <- final_results[,2:3]
  return(add_data)  
}

```

```{r functions_spatialDGEs}
## Functions related to finding spatially variable genes within a given section of the head
### Fit loess over the expression of the given genes as a function of their distance to the edge of the selected sections. returns a list containing xxx
####object: Seurat object
#### cluster: clusters to be included
#### paramdf: parameter df for sections of interest
#### genes: genes to be analysed as Gene.stable.IDs
#### gorder: order of genes for plots; if NULL, genes are sorted alphabetically, default NULL
EotM<-function(object, cluster, paramdf, genes, gorder= NULL){
  meta<- object@meta.data
  meta_HQ<- meta[meta$Brain_ID %in% paramdf$Brain_ID & meta$Sp_clusters %in% cluster, ]
  meta_HQ_right<-meta_HQ[meta_HQ$y_rot_scaled > 0,]
  meta_HQ_left<-meta_HQ[meta_HQ$y_rot_scaled < 0,]
  
  ## Do everything for the right brains
  
  right_list<-list()
  if ("right" %in% paramdf$side ){
    brains<- sort(unique(paramdf$Brain_ID[paramdf$side == "right"]))
    for (i in 1:length(brains)){
      right_list[[i]]<-list()
      right_list[[i]]$plots<-list()
      right_list[[i]]$data<-list()
      dfplot<- meta_HQ_right[meta_HQ_right$Brain_ID == brains[i],]
      paramplot<- paramdf[paramdf$side == "right" & paramdf$Brain_ID == brains[i],]
      right_list[[i]]$plots$Selection<-ggplot()+
        geom_point(data=dfplot, aes(x=y_rot_scaled, y=x_rot_scaled, colour = Sp_clusters), size=0.5)+
        geom_hline(data=paramplot, aes_string(yintercept = paramplot$limit_x_rot), size=0.5, colour = "red")+
        geom_vline(data=paramplot, aes_string(xintercept = paramplot$limit_y_rot), size=0.5, colour = "red")+
        geom_point(data=paramplot, aes_string(x=paramplot$center_y_rot, y=paramplot$center_x_rot),size=1, colour = "red", shape=8)+
        theme_bw()+
        scale_y_reverse()+
        scale_color_manual(values = cols_cl, name="Cluster")+
        #ggtitle(paste0(brains[i], " right side"))+
        coord_fixed()+
        guides(colour=F)+
        xlab("")+ylab("")
      
      right_list[[i]]$data$raw<-scapNA(dfplot, "XC")
      right_list[[i]]$data$zoom<-  right_list[[i]]$data$raw[ right_list[[i]]$data$raw$y_rot_scaled > paramplot$limit_y_rot &  right_list[[i]]$data$raw$x_rot_scaled < paramplot$limit_x_rot,]
      right_list[[i]]$data$zoom<-scapNA(right_list[[i]]$data$zoom, "XC")
      
      right_list[[i]]$data$zoom$distance_0<- sqrt((right_list[[i]]$data$zoom$x_rot_scaled-paramplot$center_x_rot)^2 + (right_list[[i]]$data$zoom$y_rot_scaled-paramplot$center_y_rot)^2)
      right_list[[i]]$data$zoom$distance_0_scaled<-right_list[[i]]$data$zoom$distance_0-min(right_list[[i]]$data$zoom$distance_0)
      
      right_list[[i]]$plots$Distance<-ggplot()+
        geom_point(data=dfplot, aes(x=y_rot_scaled, y=x_rot_scaled),colour="black", size=0.5, alpha=0.5)+
        geom_point(data=right_list[[i]]$data$zoom, aes(x=y_rot_scaled, y=x_rot_scaled, colour = distance_0*0.647), size=0.5)+
        scale_color_gradientn(colours=c("blue4","blue","lightblue","orange", "red","red4"), values = c(0,0.2,0.4,0.6,0.8,1),name = "Distance in µm")+
        geom_hline(data=paramplot, aes_string(yintercept = paramplot$limit_x_rot), size=0.5, colour = "red")+
        geom_vline(data=paramplot, aes_string(xintercept = paramplot$limit_y_rot), size=0.5, colour = "red")+
        geom_point(data=paramplot, aes_string(x=paramplot$center_y_rot, y=paramplot$center_x_rot),size=1, colour = "red", shape=8)+
        theme_bw()+
        scale_y_reverse()+
        coord_fixed()+
        xlab("")+ylab("")
      
      max_dist<- max(right_list[[i]]$data$zoom$distance_0 )
      right_list[[i]]$data$zoom$distance_edge<-max_dist-right_list[[i]]$data$zoom$distance_0  
      right_list[[i]]$plots$DistanceEdge<-ggplot()+
        geom_point(data=dfplot, aes(x=y_rot_scaled, y=x_rot_scaled),colour="black", size=0.5, alpha=0.5)+
        geom_point(data=right_list[[i]]$data$zoom, aes(x=y_rot_scaled, y=x_rot_scaled, colour = distance_edge*0.647), size=0.5)+
        scale_color_gradientn(colours=c("blue4","blue","lightblue","orange", "red","red4"), values = c(0,0.2,0.4,0.6,0.8,1),name = "Distance from edge in µm")+
        geom_hline(data=paramplot, aes_string(yintercept = paramplot$limit_x_rot), size=0.5, colour = "red")+
        geom_vline(data=paramplot, aes_string(xintercept = paramplot$limit_y_rot), size=0.5, colour = "red")+
        geom_point(data=paramplot, aes_string(x=paramplot$center_y_rot, y=paramplot$center_x_rot),size=1, colour = "red", shape=8)+
        theme_bw()+
        scale_y_reverse()+
        coord_fixed()+
        xlab("")+ylab("")
      
      right_list[[i]]$plots$Distancebox<-ggplot()+
        geom_boxplot(data=right_list[[i]]$data$zoom, aes(x=Sp_clusters, y=distance_0*0.647, colour = Sp_clusters), size=0.5)+
        geom_jitter(data=right_list[[i]]$data$zoom, aes(x=Sp_clusters, y=distance_0*0.647, colour = Sp_clusters), size=0.5)+
        theme_bw()+
        scale_color_manual(values = cols_cl, name="Cluster")+
        guides(colour = guide_legend(override.aes = list(size=5,nrow=1)))+ theme(legend.position = "top")+
        xlab("")+ylab("Distance in µm")
      
      right_list[[i]]$plots$Distancebox_edge<-ggplot()+
        geom_boxplot(data=right_list[[i]]$data$zoom, aes(x=Sp_clusters, y=distance_edge*0.647, colour = Sp_clusters), size=0.5)+
        geom_jitter(data=right_list[[i]]$data$zoom, aes(x=Sp_clusters, y=distance_edge*0.647, colour = Sp_clusters), size=0.5)+
        theme_bw()+
        scale_color_manual(values = cols_cl, name="Cluster")+
        guides(colour = guide_legend(override.aes = list(size=5,nrow=1)))+ theme(legend.position = "top")+
        xlab("")+ylab("Distance in µm")
      
      
      DefaultAssay(object) <- "SCT"
      data <- FetchData(object = object,vars = c("XC", genes))
      data<-data[data$XC %in% right_list[[i]]$data$zoom$XC,]
      data_long<- data %>%   tidyr::pivot_longer(cols = 2:ncol(data),names_to = "Gene.stable.ID",values_to = "norm_Expr")
      data_long<- dplyr::left_join(data_long, All_genes_AgamP4_PEST, by="Gene.stable.ID")
      data_long$Gene.name[is.na(data_long$Gene.name)]<- data_long$Gene.stable.ID[is.na(data_long$Gene.name)]
      
      data_long<- dplyr::left_join(data_long, right_list[[i]]$data$zoom[, c("XC", "distance_0", "distance_0_scaled", "Sp_clusters", "x_rot_scaled", "y_rot_scaled")], by="XC")
      if (!(is_null(gorder))){
        data_long$Gene.name<-factor(data_long$Gene.name, levels=c(gorder))
      }
      
      right_list[[i]]$plots$Geneplot<-ggplot()+
        geom_point(data=data_long, aes(x=distance_0*0.647, y=norm_Expr, colour = Sp_clusters), size=0.5)+
        geom_smooth(data=data_long, aes(x=distance_0*0.647, y=norm_Expr), size=1, se = F, colour="black", linetype="dashed")+
        theme_bw()+
        scale_color_manual(values = cols_cl, name="Cluster")+
        guides(colour = guide_legend(override.aes = list(size=5), ncol=2))+
        facet_wrap(~Gene.name, ncol=length(genes))+
        xlab("Distance in µm")+ylab("")+guides(colour=F)
      
      max_dist<-max(data_long$distance_0)
      data_long$distance_edge<- max_dist-data_long$distance_0 
      right_list[[i]]$plots$Geneplot_edge<-ggplot()+
        geom_point(data=data_long, aes(x=distance_edge*0.647, y=norm_Expr, colour = Sp_clusters), size=0.5)+
        geom_smooth(data=data_long, aes(x=distance_edge*0.647, y=norm_Expr), size=1, se = F, colour="black", linetype="dashed")+
        theme_bw()+
        scale_color_manual(values = cols_cl, name="Cluster")+
        guides(colour = guide_legend(override.aes = list(size=5), ncol=2))+
        facet_wrap(~Gene.name, ncol=length(genes))+
        xlab("Distance in µm")+ylab("")+guides(colour=F)
      
      right_list[[i]]$plots$Geneplot_spatial<-ggplot()+
        geom_hline(data=paramplot, aes_string(yintercept = paramplot$limit_x_rot), size=0.5, colour = "red")+
        geom_vline(data=paramplot, aes_string(xintercept = paramplot$limit_y_rot), size=0.5, colour = "red")+
        geom_point(data=paramplot, aes_string(x=paramplot$center_y_rot, y=paramplot$center_x_rot), size=1, colour = "red", shape=8)+
        geom_point(data=dfplot, aes(x=y_rot_scaled, y=x_rot_scaled),colour="black", size=0.5, alpha=0.5)+
        geom_point(data=data_long, aes(x=y_rot_scaled, y=x_rot_scaled, colour = norm_Expr), size=0.5)+
        scale_color_gradientn(colours=c("grey","orange", "red","red4"), values = c(0,0.33,0.66,1), name = "Normalised Expression")+
        theme_bw()+facet_wrap(~Gene.name, ncol=length(genes))+coord_fixed()+ xlab("")+ylab("")+scale_y_reverse()
      
      tmp<- cowplot::plot_grid(right_list[[i]]$plots$Selection , right_list[[i]]$plots$Distance, right_list[[i]]$plots$Distancebox,  ncol=3, rel_widths = c(1,1,0.6))
      right_list[[i]]$plots$CombinePLot<- cowplot::plot_grid(tmp,right_list[[i]]$plots$Geneplot_spatial, right_list[[i]]$plots$Geneplot,   nrow=3)
      
    }
    names(right_list)<-brains
  }
  
  left_list<-list()
  if ("left" %in% paramdf$side ){
    ## Do everything for the left brains
    brains<- sort(unique(paramdf$Brain_ID[paramdf$side == "left"]))
    for (i in 1:length(brains)){
      left_list[[i]]<-list()
      left_list[[i]]$plots<-list()
      left_list[[i]]$data<-list()
      dfplot<- meta_HQ_left[meta_HQ_left$Brain_ID == brains[i],]
      paramplot<- paramdf[paramdf$side == "left" & paramdf$Brain_ID == brains[i],]
      left_list[[i]]$plots$Selection<-ggplot()+
        geom_point(data=dfplot, aes(x=y_rot_scaled, y=x_rot_scaled, colour = Sp_clusters), size=0.5)+
        geom_hline(data=paramplot, aes_string(yintercept = paramplot$limit_x_rot), size=0.5, colour = "red")+
        geom_vline(data=paramplot, aes_string(xintercept = paramplot$limit_y_rot), size=0.5, colour = "red")+
        geom_point(data=paramplot, aes_string(x=paramplot$center_y_rot, y=paramplot$center_x_rot), size=1, colour = "red", shape=8)+
        scale_y_reverse()+
        theme_bw()+
        scale_color_manual(values = cols_cl, name="Cluster")+
        #ggtitle(paste0(brains[i], " left side"))+guides(colour=F)+
        xlab("")+ylab("")
      
      
      
      left_list[[i]]$data$raw<-scapNA(dfplot, "XC")
      left_list[[i]]$data$zoom<-  left_list[[i]]$data$raw[ left_list[[i]]$data$raw$y_rot_scaled < paramplot$limit_y_rot &  left_list[[i]]$data$raw$x_rot_scaled < paramplot$limit_x_rot,]
      left_list[[i]]$data$zoom<-scapNA(left_list[[i]]$data$zoom, "XC")
      
      left_list[[i]]$data$zoom$distance_0<- sqrt((left_list[[i]]$data$zoom$x_rot_scaled-paramplot$center_x_rot)^2 + (left_list[[i]]$data$zoom$y_rot_scaled-paramplot$center_y_rot)^2)
      left_list[[i]]$data$zoom$distance_0_scaled<-left_list[[i]]$data$zoom$distance_0- min(left_list[[i]]$data$zoom$distance_0)
      
      left_list[[i]]$plots$Distance<-ggplot()+
        geom_point(data=dfplot, aes(x=y_rot_scaled, y=x_rot_scaled),colour="black", size=0.5, alpha=0.5)+
        geom_point(data=left_list[[i]]$data$zoom, aes(x=y_rot_scaled, y=x_rot_scaled, colour = distance_0*0.647), size=0.5)+
        scale_color_gradientn(colours=c("blue4","blue","lightblue","orange", "red","red4"), values = c(0,0.2,0.4,0.6,0.8,1),name = "Distance in µm")+
        geom_hline(data=paramplot, aes_string(yintercept = paramplot$limit_x_rot), size=0.5, colour = "red")+
        geom_vline(data=paramplot, aes_string(xintercept = paramplot$limit_y_rot), size=0.5, colour = "red")+
        geom_point(data=paramplot, aes_string(x=paramplot$center_y_rot, y=paramplot$center_x_rot), size=1, colour = "red", shape=8)+
        scale_y_reverse()+
        theme_bw()+xlab("")+ylab("")
      
      max_dist<- max(left_list[[i]]$data$zoomdistance_0 )
      left_list[[i]]$data$zoom$distance_edge<-max_dist-left_list[[i]]$data$zoom$distance_0  
      left_list[[i]]$plots$DistanceEdge<-ggplot()+
        geom_point(data=dfplot, aes(x=y_rot_scaled, y=x_rot_scaled),colour="black", size=0.5, alpha=0.5)+
        geom_point(data=left_list[[i]]$data$zoom, aes(x=y_rot_scaled, y=x_rot_scaled, colour = distance_edge*0.647), size=0.5)+
        scale_color_gradientn(colours=c("blue4","blue","lightblue","orange", "red","red4"), values = c(0,0.2,0.4,0.6,0.8,1),name = "Distance from edge in µm")+
        geom_hline(data=paramplot, aes_string(yintercept = paramplot$limit_x_rot), size=0.5, colour = "red")+
        geom_vline(data=paramplot, aes_string(xintercept = paramplot$limit_y_rot), size=0.5, colour = "red")+
        geom_point(data=paramplot, aes_string(x=paramplot$center_y_rot, y=paramplot$center_x_rot),size=1, colour = "red", shape=8)+
        theme_bw()+
        scale_y_reverse()+
        coord_fixed()+
        xlab("")+ylab("")
      
      left_list[[i]]$plots$Distancebox<-ggplot()+
        geom_boxplot(data=left_list[[i]]$data$zoom, aes(x=Sp_clusters, y=distance_0*0.647, colour = Sp_clusters), size=0.5)+
        geom_jitter(data=left_list[[i]]$data$zoom, aes(x=Sp_clusters, y=distance_0*0.647, colour = Sp_clusters), size=0.5)+
        theme_bw()+
        scale_color_manual(values = cols_cl, name="Cluster")+
        guides(colour = guide_legend(override.aes = list(size=5,nrow=1)))+ theme(legend.position = "top")+
        xlab("Cluster")+ylab("Distance in µm")
      
      left_list[[i]]$plots$Distancebox_edge<-ggplot()+
        geom_boxplot(data=left_list[[i]]$data$zoom, aes(x=Sp_clusters, y=distance_edge*0.647, colour = Sp_clusters), size=0.5)+
        geom_jitter(data=left_list[[i]]$data$zoom, aes(x=Sp_clusters, y=distance_edge*0.647, colour = Sp_clusters), size=0.5)+
        theme_bw()+
        scale_color_manual(values = cols_cl, name="Cluster")+
        guides(colour = guide_legend(override.aes = list(size=5,nrow=1)))+ theme(legend.position = "top")+
        xlab("Cluster")+ylab("Distance in µm")
      
      DefaultAssay(object) <- "SCT"
      data <- FetchData(object = object,vars = c("XC", genes))
      data<-data[data$XC %in% left_list[[i]]$data$zoom$XC,]
      data_long<- data %>%   tidyr::pivot_longer(cols = 2:ncol(data),names_to = "Gene.stable.ID",values_to = "norm_Expr")
      data_long<- dplyr::left_join(data_long, All_genes_AgamP4_PEST, by="Gene.stable.ID")
      data_long$Gene.name[is.na(data_long$Gene.name)]<- data_long$Gene.stable.ID[is.na(data_long$Gene.name)]
      
      data_long<- dplyr::left_join(data_long, left_list[[i]]$data$zoom[, c("XC", "distance_0", "distance_0_scaled","Sp_clusters","x_rot_scaled" ,"y_rot_scaled")], by="XC")
      if (!(is_null(gorder))){
        data_long$Gene.name<-factor(data_long$Gene.name, levels=c(gorder))
      }
      
      left_list[[i]]$plots$Geneplot<-ggplot()+
        geom_point(data=data_long, aes(x=distance_0*0.647, y=norm_Expr, colour = Sp_clusters), size=0.5)+
        geom_smooth(data=data_long, aes(x=distance_0*0.647, y=norm_Expr), size=1, se = F, colour="black", linetype="dashed")+
        theme_bw()+
        scale_color_manual(values = cols_cl, name="Cluster")+
        guides(colour = guide_legend(override.aes = list(size=5), ncol=2))+
        facet_wrap(~Gene.name, ncol=length(genes))+
        xlab("Distance in µm")+ylab("")+guides(colour=F)
      
      max_dist<-max(data_long$distance_0)
      data_long$distance_edge<- max_dist-data_long$distance_0 
      left_list[[i]]$plots$Geneplot_edge<-ggplot()+
        geom_point(data=data_long, aes(x=distance_edge*0.647, y=norm_Expr, colour = Sp_clusters), size=0.5)+
        geom_smooth(data=data_long, aes(x=distance_edge*0.647, y=norm_Expr), size=1, se = F, colour="black", linetype="dashed")+
        theme_bw()+
        scale_color_manual(values = cols_cl, name="Cluster")+
        guides(colour = guide_legend(override.aes = list(size=5), ncol=2))+
        facet_wrap(~Gene.name, ncol=length(genes))+
        xlab("Distance in µm")+ylab("")+guides(colour=F)
      
      left_list[[i]]$plots$Geneplot_spatial<-ggplot()+
        geom_hline(data=paramplot, aes_string(yintercept = paramplot$limit_x_rot), size=0.5, colour = "red")+
        geom_vline(data=paramplot, aes_string(xintercept = paramplot$limit_y_rot), size=0.5, colour = "red")+
        geom_point(data=paramplot, aes_string(x=paramplot$center_y_rot, y=paramplot$center_x_rot), size=1, colour = "red", shape=8)+
        geom_point(data=dfplot, aes(x=y_rot_scaled, y=x_rot_scaled),colour="black", size=0.5, alpha=0.5)+
        geom_point(data=data_long, aes(x=y_rot_scaled, y=x_rot_scaled, colour = norm_Expr), size=0.5)+
        scale_color_gradientn(colours=c("grey","orange", "red","red4"), values = c(0,0.33,0.66,1), name = "Normalised Expression")+
        theme_bw()+facet_wrap(~Gene.name, ncol=length(genes))+coord_fixed()+xlab("")+ylab("")+
        scale_y_reverse()
      
      tmp<- cowplot::plot_grid(left_list[[i]]$plots$Selection , left_list[[i]]$plots$Distance, left_list[[i]]$plots$Distancebox, ncol=3, rel_widths = c(1,1,0.6))
      left_list[[i]]$plots$CombinePLot<- cowplot::plot_grid(tmp, left_list[[i]]$plots$Geneplot_spatial, left_list[[i]]$plots$Geneplot, nrow=2)
      
      
    }
    names(left_list)<-brains
  }
  results<- list(right_list, left_list)
  names(results)<-c("right", "left")
  return(results)
}

### Fit natural cubic spline fit over the expression of the given genes as a function of their distance to the edge of the selected sections. returns a list containing xxx
####object: Seurat object
#### cluster: clusters to be included
#### paramdf: parameter df for sections of interest
#### genes: genes to be analysed as Gene.stable.IDs
#### freedom: degrees of freedom to be used in natural cubic spline fit, default 10
EotM_fit<-function(object, cluster, paramdf, genes, freedom=10){
  meta<- object@meta.data
  meta_HQ<- meta[meta$Brain_ID %in% paramdf$Brain_ID & meta$Sp_clusters %in% cluster, ]
  meta_HQ_right<-meta_HQ[meta_HQ$y_rot_scaled > 0,]
  meta_HQ_left<-meta_HQ[meta_HQ$y_rot_scaled < 0,]
  
  ## if section is on the right side of the brain
  
  right_list<-list()
  if ("right" %in% paramdf$side ){
    brains<- sort(unique(paramdf$Brain_ID[paramdf$side == "right"]))
    for (i in 1:length(brains)){
      right_list[[i]]<-list()
      right_list[[i]]$plots<-list()
      right_list[[i]]$data<-list()
      dfplot<- meta_HQ_right[meta_HQ_right$Brain_ID == brains[i],]
      paramplot<- filter(paramdf, side == "right" & Brain_ID == brains[i])
      right_list[[i]]$plots$Selection<-ggplot()+
        geom_point(data=dfplot, aes(x=y_rot_scaled, y=x_rot_scaled, colour = Sp_clusters), size=0.5)+
        geom_hline(data=paramplot, aes_string(yintercept = paramplot$limit_x_rot), size=0.5, colour = "red")+
        geom_vline(data=paramplot, aes_string(xintercept = paramplot$limit_y_rot), size=0.5, colour = "red")+
        geom_point(data=paramplot, aes_string(x=paramplot$center_y_rot, y=paramplot$center_x_rot),size=1, colour = "red", shape=8)+
        theme_bw()+
        scale_y_reverse()+
        scale_color_manual(values = cols_cl, name="Cluster")+
        #ggtitle(paste0(brains[i], " right side"))+
        coord_fixed()+
        guides(colour=F)+
        xlab("")+ylab("")
      
      right_list[[i]]$data$raw<-dfplot
      right_list[[i]]$data$zoom<-  filter(right_list[[i]]$data$raw, y_rot_scaled > paramplot$limit_y_rot &  x_rot_scaled < paramplot$limit_x_rot)
      
      right_list[[i]]$data$zoom$distance_0<- sqrt((right_list[[i]]$data$zoom$x_rot_scaled-paramplot$center_x_rot)^2 + (right_list[[i]]$data$zoom$y_rot_scaled-paramplot$center_y_rot)^2)
      right_list[[i]]$data$zoom$distance_0_scaled<-right_list[[i]]$data$zoom$distance_0-min(right_list[[i]]$data$zoom$distance_0)
      
      right_list[[i]]$plots$Distance<-ggplot()+
        geom_point(data=dfplot, aes(x=y_rot_scaled, y=x_rot_scaled),colour="black", size=0.5, alpha=0.5)+
        geom_point(data=right_list[[i]]$data$zoom, aes(x=y_rot_scaled, y=x_rot_scaled, colour = distance_0*0.647), size=0.5)+
        scale_color_gradientn(colours=c("blue4","blue","lightblue","orange", "red","red4"), values = c(0,0.2,0.4,0.6,0.8,1),name = "Distance in µm")+
        geom_hline(data=paramplot, aes_string(yintercept = paramplot$limit_x_rot), size=0.5, colour = "red")+
        geom_vline(data=paramplot, aes_string(xintercept = paramplot$limit_y_rot), size=0.5, colour = "red")+
        geom_point(data=paramplot, aes_string(x=paramplot$center_y_rot, y=paramplot$center_x_rot),size=1, colour = "red", shape=8)+
        theme_bw()+
        scale_y_reverse()+
        coord_fixed()+
        xlab("")+ylab("")
      
      right_list[[i]]$plots$Distancebox<-ggplot()+
        geom_boxplot(data=right_list[[i]]$data$zoom, aes(x=Sp_clusters, y=distance_0*0.647, colour = Sp_clusters), size=0.5)+
        geom_jitter(data=right_list[[i]]$data$zoom, aes(x=Sp_clusters, y=distance_0*0.647, colour = Sp_clusters), size=0.5)+
        theme_bw()+
        scale_color_manual(values = cols_cl, name="Cluster")+
        guides(colour = guide_legend(override.aes = list(size=5), ncol=2))+
        xlab("")+ylab("Distance in µm")
      
      ### Extract genes from SCT data matrix
      DefaultAssay(object) <- "SCT"
      data <- FetchData(object = object,vars = c("XC", genes))
      data<-data[data$XC %in% right_list[[i]]$data$zoom$XC,]
      data_long<- data %>%   tidyr::pivot_longer(cols = 2:ncol(data),names_to = "Gene.stable.ID",values_to = "norm_Expr")
      data_long<- dplyr::left_join(data_long, All_genes_AgamP4_PEST, by="Gene.stable.ID")
      data_long$Gene.name[is.na(data_long$Gene.name)]<- data_long$Gene.stable.ID[is.na(data_long$Gene.name)]
      
      data_long<- dplyr::left_join(data_long, right_list[[i]]$data$zoom[, c("XC", "distance_0", "distance_0_scaled", "Sp_clusters", "x_rot_scaled", "y_rot_scaled")], by="XC")
      right_list[[i]]$data$ExpressionData<-data_long
      ### Perform fit per gene and extract expression range(max/min, mean, meadian)
      library(purrr)
      library(splines)
      right_list[[i]]$PolynomialFits <- data_long %>% 
        split(f = .$Gene.stable.ID) %>% 
        map(~stats::lm(norm_Expr ~ poly(distance_0,freedom,raw=TRUE), data=.x))
      
      right_list[[i]]$NatSplineFits <- data_long %>% 
        split(f = .$Gene.stable.ID) %>% 
        map(~stats::lm(norm_Expr ~ ns(distance_0,freedom), data=.x))
      
      right_list[[i]]$BSplineFits <- data_long %>% 
        split(f = .$Gene.stable.ID) %>% 
        map(~stats::lm(norm_Expr ~ bs(distance_0,freedom), data=.x))

      right_list[[i]]$data$expr_ranges<- data_long %>% group_by(Gene.stable.ID) %>% summarise("Max"= max(norm_Expr),"Min"= min(norm_Expr),"Median"= median(norm_Expr),"Mean"= mean(norm_Expr)) %>% as.data.frame()
     
      right_list[[i]]$plots$min_max<-ggplot()+
     geom_point(data=right_list[[i]]$data$expr_ranges, aes(x=Min, y=Max))+
    theme_bw()+ expand_limits(x = 0, y = 0)
      
      right_list[[i]]$plots$median_max<-ggplot()+
        geom_point(data=right_list[[i]]$data$expr_ranges, aes(x=Median, y=Max))+
        theme_bw()+ expand_limits(x = 0, y = 0)
      
      right_list[[i]]$plots$mean_max<-ggplot()+
        geom_point(data=right_list[[i]]$data$expr_ranges, aes(x=Mean, y=Max))+
        theme_bw()+ expand_limits(x = 0, y = 0)
  

      right_list[[i]]$data$Polynomial_stats <- plyr::ldply(right_list[[i]]$PolynomialFits, function(x){ data.frame(r_sq = summary(x)$r.squared, adj.r_squ = summary(x)$adj.r.squared)})
      right_list[[i]]$data$NatSpline_stats <- plyr::ldply(right_list[[i]]$NatSplineFits, function(x){ data.frame(r_sq = summary(x)$r.squared, adj.r_squ = summary(x)$adj.r.squared)})
      right_list[[i]]$data$BSpline_stats <- plyr::ldply(right_list[[i]]$BSplineFits, function(x){ data.frame(r_sq = summary(x)$r.squared, adj.r_squ = summary(x)$adj.r.squared)})
      
      right_list[[i]]$data$combined_stats<- gdata::combine(right_list[[i]]$data$Polynomial_stats, right_list[[i]]$data$NatSpline_stats,right_list[[i]]$data$BSpline_stats)
      right_list[[i]]$data$combined_stats$source<- sapply(as.character(right_list[[i]]$data$combined_stats$source), function(x){ strsplit(strsplit(x, split = "$", fixed=T)[[1]][3], split = "_")[[1]][1]})
      
      
      right_list[[i]]$plots$Rsqu_distr<-ggplot()+
        geom_density(data=right_list[[i]]$data$combined_stats, aes(x=r_sq, colour=source))+
        theme_bw()+xlim(0,1)
      
      right_list[[i]]$plots$adj.Rsqu_distr<-ggplot()+
        geom_density(data=right_list[[i]]$data$combined_stats, aes(x=adj.r_squ,colour=source))+
        theme_bw()+xlim(0,1)
    }
    names(right_list)<-brains
  }
  
  ### If section is on the left side of the brain
  left_list<-list()
  if ("left" %in% paramdf$side ){
    brains<- sort(unique(paramdf$Brain_ID[paramdf$side == "left"]))
    for (i in 1:length(brains)){
      left_list[[i]]<-list()
      left_list[[i]]$plots<-list()
      left_list[[i]]$data<-list()
      dfplot<- meta_HQ_left[meta_HQ_left$Brain_ID == brains[i],]
      paramplot<- filter(paramdf, side == "left" & Brain_ID == brains[i])
      left_list[[i]]$plots$Selection<-ggplot()+
        geom_point(data=dfplot, aes(x=y_rot_scaled, y=x_rot_scaled, colour = Sp_clusters), size=0.5)+
        geom_hline(data=paramplot, aes_string(yintercept = paramplot$limit_x_rot), size=0.5, colour = "red")+
        geom_vline(data=paramplot, aes_string(xintercept = paramplot$limit_y_rot), size=0.5, colour = "red")+
        geom_point(data=paramplot, aes_string(x=paramplot$center_y_rot, y=paramplot$center_x_rot),size=1, colour = "red", shape=8)+
        theme_bw()+
        scale_y_reverse()+
        scale_color_manual(values = cols_cl, name="Cluster")+
        #ggtitle(paste0(brains[i], " left side"))+
        coord_fixed()+
        guides(colour=F)+
        xlab("")+ylab("")
      
      left_list[[i]]$data$raw<-dfplot
      left_list[[i]]$data$zoom<-  filter(left_list[[i]]$data$raw, y_rot_scaled > paramplot$limit_y_rot &  x_rot_scaled < paramplot$limit_x_rot)
      
      left_list[[i]]$data$zoom$distance_0<- sqrt((left_list[[i]]$data$zoom$x_rot_scaled-paramplot$center_x_rot)^2 + (left_list[[i]]$data$zoom$y_rot_scaled-paramplot$center_y_rot)^2)
      left_list[[i]]$data$zoom$distance_0_scaled<-left_list[[i]]$data$zoom$distance_0-min(left_list[[i]]$data$zoom$distance_0)
      
      left_list[[i]]$plots$Distance<-ggplot()+
        geom_point(data=dfplot, aes(x=y_rot_scaled, y=x_rot_scaled),colour="black", size=0.5, alpha=0.5)+
        geom_point(data=left_list[[i]]$data$zoom, aes(x=y_rot_scaled, y=x_rot_scaled, colour = distance_0*0.647), size=0.5)+
        scale_color_gradientn(colours=c("blue4","blue","lightblue","orange", "red","red4"), values = c(0,0.2,0.4,0.6,0.8,1),name = "Distance in µm")+
        geom_hline(data=paramplot, aes_string(yintercept = paramplot$limit_x_rot), size=0.5, colour = "red")+
        geom_vline(data=paramplot, aes_string(xintercept = paramplot$limit_y_rot), size=0.5, colour = "red")+
        geom_point(data=paramplot, aes_string(x=paramplot$center_y_rot, y=paramplot$center_x_rot),size=1, colour = "red", shape=8)+
        theme_bw()+
        scale_y_reverse()+
        coord_fixed()+
        xlab("")+ylab("")
      
      left_list[[i]]$plots$Distancebox<-ggplot()+
        geom_boxplot(data=left_list[[i]]$data$zoom, aes(x=Sp_clusters, y=distance_0*0.647, colour = Sp_clusters), size=0.5)+
        geom_jitter(data=left_list[[i]]$data$zoom, aes(x=Sp_clusters, y=distance_0*0.647, colour = Sp_clusters), size=0.5)+
        theme_bw()+
        scale_color_manual(values = cols_cl, name="Cluster")+
        guides(colour = guide_legend(override.aes = list(size=5), ncol=2))+
        xlab("")+ylab("Distance in µm")
      
      ### Extract genes from SCT data matrix
      DefaultAssay(object) <- "SCT"
      data <- FetchData(object = object,vars = c("XC", genes))
      data<-data[data$XC %in% left_list[[i]]$data$zoom$XC,]
      data_long<- data %>%   tidyr::pivot_longer(cols = 2:ncol(data),names_to = "Gene.stable.ID",values_to = "norm_Expr")
      data_long<- dplyr::left_join(data_long, All_genes_AgamP4_PEST, by="Gene.stable.ID")
      data_long$Gene.name[is.na(data_long$Gene.name)]<- data_long$Gene.stable.ID[is.na(data_long$Gene.name)]
      
      data_long<- dplyr::left_join(data_long, left_list[[i]]$data$zoom[, c("XC", "distance_0", "distance_0_scaled", "Sp_clusters", "x_rot_scaled", "y_rot_scaled")], by="XC")
      left_list[[i]]$data$ExpressionData<-data_long
      ### Perform fit per gene and extract expression range(max/min, mean, meadian)
      library(purrr)
      library(splines)
      left_list[[i]]$PolynomialFits <- data_long %>% 
        split(f = .$Gene.stable.ID) %>% 
        map(~stats::lm(norm_Expr ~ poly(distance_0,freedom,raw=TRUE), data=.x))
      
      left_list[[i]]$NatSplineFits <- data_long %>% 
        split(f = .$Gene.stable.ID) %>% 
        map(~stats::lm(norm_Expr ~ ns(distance_0,freedom), data=.x))
      
      left_list[[i]]$BSplineFits <- data_long %>% 
        split(f = .$Gene.stable.ID) %>% 
        map(~stats::lm(norm_Expr ~ bs(distance_0,freedom), data=.x))
      
      left_list[[i]]$data$expr_ranges<- data_long %>% group_by(Gene.stable.ID) %>% summarise("Max"= max(norm_Expr),"Min"= min(norm_Expr),"Median"= median(norm_Expr),"Mean"= mean(norm_Expr)) %>% as.data.frame()
      
      left_list[[i]]$plots$min_max<-ggplot()+
        geom_point(data=left_list[[i]]$data$expr_ranges, aes(x=Min, y=Max))+
        theme_bw()+ expand_limits(x = 0, y = 0)
      
      left_list[[i]]$plots$median_max<-ggplot()+
        geom_point(data=left_list[[i]]$data$expr_ranges, aes(x=Median, y=Max))+
        theme_bw()+ expand_limits(x = 0, y = 0)
      
      left_list[[i]]$plots$mean_max<-ggplot()+
        geom_point(data=left_list[[i]]$data$expr_ranges, aes(x=Mean, y=Max))+
        theme_bw()+ expand_limits(x = 0, y = 0)
      
      
      left_list[[i]]$data$Polynomial_stats <- plyr::ldply(left_list[[i]]$PolynomialFits, function(x){ data.frame(r_sq = summary(x)$r.squared, adj.r_squ = summary(x)$adj.r.squared)})
      left_list[[i]]$data$NatSpline_stats <- plyr::ldply(left_list[[i]]$NatSplineFits, function(x){ data.frame(r_sq = summary(x)$r.squared, adj.r_squ = summary(x)$adj.r.squared)})
      left_list[[i]]$data$BSpline_stats <- plyr::ldply(left_list[[i]]$BSplineFits, function(x){ data.frame(r_sq = summary(x)$r.squared, adj.r_squ = summary(x)$adj.r.squared)})
      
      left_list[[i]]$data$combined_stats<- gdata::combine(left_list[[i]]$data$Polynomial_stats, left_list[[i]]$data$NatSpline_stats,left_list[[i]]$data$BSpline_stats)
      left_list[[i]]$data$combined_stats$source<- sapply(as.character(left_list[[i]]$data$combined_stats$source), function(x){ strsplit(strsplit(x, split = "$", fixed=T)[[1]][3], split = "_")[[1]][1]})
      
      
      left_list[[i]]$plots$Rsqu_distr<-ggplot()+
        geom_density(data=left_list[[i]]$data$combined_stats, aes(x=r_sq, colour=source))+
        theme_bw()+xlim(0,1)
      
      left_list[[i]]$plots$adj.Rsqu_distr<-ggplot()+
        geom_density(data=left_list[[i]]$data$combined_stats, aes(x=adj.r_squ,colour=source))+
        theme_bw()+xlim(0,1)
    }
    names(left_list)<-brains
  }
  results<- list(right_list, left_list)
  names(results)<-c("right", "left")
  return(results)
}
```

```{r functions_DGE}
## Functions related to Differential gene expression
### Find all markers for specific group in seurat object, i.e. FindAllMarkers function from suerat with additional plotting and filtering steps. Returns a list of Dataframes namely "AllTests" with no thresholds , "Significant" with pvalue and logfld threshold  and "Filtered" with pvalue, logFC and pct thresholds. 
#### object: Seurat Object
#### method: DGE method ,e.g. "wilcox" or "mast", default "wilcox", see "test.use" in Seurat::FindAllMarkers()
#### lf.th: logfoldchange threshold for genes to be considered differentially expressed, default 0.5
#### p.th: adjusted p value threshold for genes to be considered differentially expressed, default 0.05
#### pct.th: percent of cells expressiing the gene to be considered differentially expressed, default 0.25
#### group: group to perform DGE on, default= "ident", , see "group.by" in Seurat::FindAllMarkers()
GetAllMarkers<- function(object, method="wilcox", lf.th= 0.5, p.th=0.05, pct.th=0.25, group="ident"){
  DEG_List<-list()
  DEG_List$AllTests<-FindAllMarkers(object, logfc.threshold = 0, min.pct = 0, test.use = method, group.by = group)
  colnames(DEG_List$AllTests)[ncol(DEG_List$AllTests)]<- "Gene.stable.ID"
  
  DEG_List$AllTests<- dplyr::left_join(DEG_List$AllTests, All_genes_AgamP4_PEST, by="Gene.stable.ID")
  DEG_List$AllTests$Gene.name[is.na(DEG_List$AllTests$Gene.name)]<- DEG_List$AllTests$Gene.stable.ID[is.na(DEG_List$AllTests$Gene.name)]
  
  DEG_List$Significant<-DEG_List$AllTests[DEG_List$AllTests$p_val_adj < p.th & abs(DEG_List$AllTests$avg_log2FC) > lf.th, ]
  DEG_List$Filtered<-DEG_List$Significant[DEG_List$Significant$pct.1 > pct.th | DEG_List$Significant$pct.2 > pct.th  , ]
 
  return(DEG_List)
}

```

```{r functions_other}
### Remove lines with NA values in data frame
#### data: dataframe
#### colname: column name to where to look for NAs
scapNA<- function(data, colname){
  data<-data[!(is.na(data[, colname])), ]
  return(data)
}
```

```{r colour_scales}
# Colours for 15 clusters, picked from paper
cols_cl<-c("#9FB6D2","#AD5729", "#FDE67E","#376964","#D4872F","#616F94","#7B9B70", "#538CB8","#D4B435", "#ABDB9D","#2F3769","#733B30", "#CE7C51", "#C79736")
names(cols_cl)<-c("0", "1","2", "3","4", "5","6", "7","8", "9","10","11","12","13")

cols_subcl<-c("#FADE0A", "#FA0A7E","#4F4C38","#42737A","#0ADAFA", "#3646FF", "#A69A3D", "#1ED274")
names(cols_subcl)<-c("cl2_0", "cl2_1", "cl2_2", "cl2_3", "cl2_4", "cl2_5", "cl2_6", "cl2_7")

# color gradient for cell type assignment heatmaps
col_br = colorRamp2(c(0,0.25, 0.5,0.75,1), c("grey95","gold1","orange2", 'red','red4'))
```

# Step 2: load data and annotation files
```{r load_annots}
## Load gene.name to Gene.stable.ID annotation data frame
All_genes_AgamP4_PEST <- read.csv(file.path(projDir, "/Data/Annotations/All_genes_AgamP4_PEST.csv"))

## Load markers for single cell clusters
An_Markers <- read.csv(file.path(projDir, "/Data/SingleCell/An_Markers.csv"))

An_Markers$cluster_annot<-rep(c("Neural progenitors (eye)", "An. cluster 1", "adPN, IPN, OPN", "Neural progenitors", "An. cluster 4", "An. cluster 5", "Pm1/Pm2, Pm4", "Octopaminergic", "An. cluster 8", "An. cluster 9", "Ensheathing glia", "L1, L2, L3, LNvs", "Optic lobe cells (generic type 1)", "An. cluster 13", "G-KC, A/B-KC", "Photoreceptors", "Tm1/TmY8", "Optic lobe cells (generic type 2)", "CCAP/circadian clock neurons", "L2", "T3", "An. cluster 21", "An. cluster 22", "Astrocyte-like glia", "L4/L5, Tm9", "Surface glia", "An. cluster 26", "An. cluster 27", "A/B*-KC", "An. cluster 29", "L3, LNvs", "A/B-KC", "G-KC, A/B-KC", "An. cluster 33", "T4/T5", "Optic lobe cells (generic type 3)", "L1", "CCAP/circadian clock neurons", "Oenocytoids", "C2, C3", "Pm1/Pm2, Pm1/Pm2/Pm3", "Poxn neurons", "Granulocytes", "An. cluster 43", "An. cluster 44"), each=20)

An_Markers$sc_cluster<-rep(c("An0","An1","An2","An3","An4","An5","An6","An7","An8","An9","An10","An11","An12","An13","An14","An15","An16","An17","An18","An19","An20","An21","An22","An23","An24","An25","An26","An27","An28","An29","An30","An31","An32","An33","An34","An35","An36","An37","An38","An39","An40","An41","An42","An43","An44"), each=20)
```

```{r load_scExpressionData}
ref_sc<-readRDS(file.path(projDir, "/Data/SingleCell/atlas_dataset.rds"))
```

```{r load_SpatialExpressionData}
## load Seurat file created by curio pipeline for bloc2 sample
rds_bloc2<- readRDS(file.path(projDir, "/Data/Spatial/bloc2_seurat.rds"))

## load Seurat file created by curio pipeline for bloc4 sample
rds_bloc4<- readRDS(file.path(projDir, "/Data/Spatial/bloc4_seurat.rds"))
```

```{r data_combine}

### Combine the data of bloc2 and bloc4
## Prepare raw UMI matrix for bloc2
counts_bloc2<- as.data.frame(as.matrix(rds_bloc2@assays$RNA@counts)) ### obtain UMI matrix from seurat object
colnames(counts_bloc2)<-paste0("bloc2_",colnames(counts_bloc2))       ### rename barcodes
rownames(counts_bloc2)<-sapply(rownames(counts_bloc2), function(x){strsplit(x, split = ":")[[1]][2]}) ## fix gene names
counts_bloc2$Genes<-rownames(counts_bloc2)
coords_bloc2<- rds_bloc2@images$slice1@coordinates    ### extract spot coordinates
rownames(coords_bloc2)<- paste0("bloc2_",rownames(coords_bloc2))
coords_tmp_bloc2<-coords_bloc2
coords_tmp_bloc2$XC<-rownames(coords_tmp_bloc2)

### create new meta data file
meta_bloc2<- data.frame(BC=rownames(rds_bloc2@meta.data), 
                        XC=paste0("bloc2_",rownames(rds_bloc2@meta.data)),
                        bloc="bloc2",
                        percent.mt=rds_bloc2@meta.data[, "percent.mt"])

meta_bloc2<- dplyr::left_join(meta_bloc2, coords_tmp_bloc2, by="XC")
rm(coords_tmp_bloc2)
rownames(meta_bloc2)<- meta_bloc2$XC

## Prepare raw UMI matrix for bloc4
counts_bloc4<- as.data.frame(as.matrix(rds_bloc4@assays$RNA@counts)) ### obtain UMI matrix from seurat object
colnames(counts_bloc4)<-paste0("bloc4_",colnames(counts_bloc4))      ### rename barcodes
rownames(counts_bloc4)<-sapply(rownames(counts_bloc4), function(x){strsplit(x, split = ":")[[1]][2]})  ## fix gene names
counts_bloc4$Genes<-rownames(counts_bloc4)
coords_bloc4<- rds_bloc4@images$slice1@coordinates                 ### extract spot coordinates
rownames(coords_bloc4)<- paste0("bloc4_",rownames(coords_bloc4))
coords_tmp_bloc4<-coords_bloc4
coords_tmp_bloc4$XC<-rownames(coords_tmp_bloc4)

### create new meta data file
meta_bloc4<- data.frame(BC=rownames(rds_bloc4@meta.data), 
                        XC=paste0("bloc4_",rownames(rds_bloc4@meta.data)),
                        bloc="bloc4",
                        percent.mt=rds_bloc4@meta.data[, "percent.mt"])

meta_bloc4<- dplyr::left_join(meta_bloc4, coords_tmp_bloc4, by="XC")
rm(coords_tmp_bloc4)
rownames(meta_bloc4)<- meta_bloc4$XC

## Combine Data
### Create combined UMI count matrix
counts<- full_join(counts_bloc2,counts_bloc4,  by="Genes")
counts<-column_to_rownames(counts, var = "Genes")
counts<-as.matrix(counts)
counts[is.na(counts)]<- 0

### Create combine meta data file
meta<- rbind(meta_bloc2,  meta_bloc4)
coords<-rbind(coords_bloc2,  coords_bloc4)

### Create new Seurat file
combined<-CreateSeuratObject(counts=counts, meta.data = meta, project = "SpatialTranscriptomics")
meta_combined<-combined@meta.data

```

# Step 3: Perform QC
```{r QC}
## Set thresholds for UMI, genes and percent.mt
ncount_th<-100
nfeat_th<-50
percmt_th<-0.4
## Subset seurat file
combined_filt <- subset(combined, subset = nCount_RNA > ncount_th & nFeature_RNA > nfeat_th & percent.mt < percmt_th)
```

# Step 4: Assign spots to individual brains 
```{r Brain_Assign, eval = FALSE}
## Code for manually assigning Spots to single brains
## bloc2
df<-combined@meta.data[combined@meta.data$bloc == "bloc2", c("x", "y")]
plot(df, col = "red", pch = 16, cex=0.1)
bloc2_1 <- gatepoints::fhs(df)
saveRDS(bloc2_1, "/Brain_Selection/bloc2_1.rds")

plot(df, col = "red", pch = 16, cex=0.1)
bloc2_2 <- gatepoints::fhs(df)
saveRDS(bloc2_2, "/Brain_Selection/bloc2_2.rds")

plot(df, col = "red", pch = 16, cex=0.1)
bloc2_3 <- gatepoints::fhs(df)
saveRDS(bloc2_3, "/Brain_Selection/bloc2_3.rds")

plot(df, col = "red", pch = 16, cex=0.1)
bloc2_4 <- gatepoints::fhs(df)
saveRDS(bloc2_4, "/Brain_Selection/bloc2_4.rds")

## bloc4
df<-combined@meta.data[combined@meta.data$bloc == "bloc2", c("x", "y")]
plot(df, col = "red", pch = 16, cex=0.1)
bloc2_1 <- gatepoints::fhs(df)
saveRDS(bloc4_1, "/Brain_Selection/bloc4_1.rds")

plot(df, col = "red", pch = 16, cex=0.1)
bloc4_2 <- gatepoints::fhs(df)
saveRDS(bloc4_2, "/Brain_Selection/bloc4_2.rds")

plot(df, col = "red", pch = 16, cex=0.1)
bloc4_3 <- gatepoints::fhs(df)
saveRDS(bloc4_3, "/Brain_Selection/bloc4_3.rds")

plot(df, col = "red", pch = 16, cex=0.1)
bloc4_4 <- gatepoints::fhs(df)
saveRDS(bloc4_4, "/Brain_Selection/bloc4_4.rds")
```

```{r Brain_Assign_load}
### Assignemnts was performed manually as depicted in the commented out code above.  Assignments are loaded directly in this script.
singlebrains<-list.files(paste0(projDir,"/Data/Spatial/Brain_Selection/"), full.names = T, pattern = "bloc2|bloc4")
sb_names<-sapply(list.files(paste0(projDir,"/Data/Spatial/Brain_Selection/"), full.names = F, pattern = "bloc2|bloc4"), function(x){strsplit(x, split=".", fixed = T)[[1]][1]})


singlebrains_list<-list()
for (i in 1:length(singlebrains)){
  singlebrains_list[[i]]<-as.data.frame(readRDS(singlebrains[i]))
}
names(singlebrains_list)<-sb_names

singlebrains_df<- bind_rows(singlebrains_list, .id = "Brain_ID")
colnames(singlebrains_df)<- c("Brain_ID", "XC")
```

```{r Brain_Rotate}
### Coordinates of single brains will be rotated around the center for visualization purposes
Rotation_df<- data.frame(Brain_ID= unique(singlebrains_df$Brain_ID),
                         Rotation_angle= c(-130,130,-120,-10,-160,45,0,-70)) ### define rotation angles

singlebrains_df<- dplyr::left_join(singlebrains_df, Rotation_df, by="Brain_ID")

Rotation_info<- combined_filt@meta.data[, c("XC", "x", "y")]                ### extract original coordinates
Rotation_info<- dplyr::left_join(Rotation_info, singlebrains_df, by="XC")

### rotate spot coordinates per brain based on defined rotation angles
Rotation_info_2<-data.frame()
for (i in unique(Rotation_info$Brain_ID)){
  tmp_df<- drop_na(Rotation_info[Rotation_info$Brain_ID == i,])
  tmp_rot<-unique(tmp_df$Rotation_angle)
  tmp_rot_df<-   artpack::rotator(tmp_df, x, y, angle = tmp_rot, anchor = "center", drop = FALSE) 
  Rotation_info_2<-rbind(Rotation_info_2,tmp_rot_df )
}

colnames(Rotation_info_2)<-c("x_rot","y_rot","XC","Brain_ID","Rotation_angle")
rownames(Rotation_info_2)<- Rotation_info_2$XC

Rotation_info_2<- Rotation_info_2[,c("Brain_ID","Rotation_angle","x_rot","y_rot")]

### Add new rotated coordinates as meta data to seurat object
combined_filt <- AddMetaData(combined_filt, metadata = Rotation_info_2)

```

# Step 5: Identify high quality brains
```{r ID_HQ}
Brains<- sort(unique(combined_filt@meta.data$Brain_ID))
HQ_df<-data.frame(Brain_ID=Brains, HQ_Brain=c("Yes", "Yes","No", "Yes", "No","No","Yes","Yes")) 

add_meta<-combined_filt@meta.data[, c("XC", "Brain_ID")]
add_meta<- dplyr::left_join(add_meta, HQ_df, by="Brain_ID")
add_HQ<- add_meta$HQ_Brain
names(add_HQ)<-add_meta$XC

combined_filt<- AddMetaData(combined_filt,add_HQ, col.name = "HQ_Brain")

HQ_Brains<- HQ_df$Brain_ID[HQ_df$HQ_Brain == "Yes"]
```

```{r HQ_scale}
vabe_list<- list()
for (i in unique(HQ_Brains)){
  vabe<-filter(combined_filt@meta.data, Brain_ID == i)
  
  maxi_x<-max(vabe$x_rot)
  maxi_y<-max(vabe$y_rot)
  
  mini_x<-min(vabe$x_rot)
  mini_y<-min(vabe$y_rot)
  
  cent_x<- (maxi_x-mini_x)/2
  cent_y<- (maxi_y-mini_y)/2
  
  vabe$x_rot_corr<- vabe$x_rot - mini_x
  vabe$y_rot_corr<- vabe$y_rot - mini_y
  
  
  corr_maxi_x<-max(vabe$x_rot_corr)
  corr_maxi_y<-max(vabe$y_rot_corr)
  
  corr_cent_x<- corr_maxi_x/2
  corr_cent_y<- corr_maxi_y/2
  
  vabe$x_rot_scaled<- vabe$x_rot_corr - corr_cent_x
  vabe$y_rot_scaled<- vabe$y_rot_corr - corr_cent_y
  
  vabe_list[[i]]<- vabe[, c("XC", "x_rot_scaled", "y_rot_scaled")]
  
}

vabe<- bind_rows(vabe_list)
add_meta<- vabe
add_meta<- add_meta[, 2:3]
rownames(add_meta)<- vabe$XC

combined_filt<- AddMetaData(combined_filt, add_meta)
```


# Step 6: Deconvolution using RCTD
```{r RCTD_run, eval = FALSE}
ref_counts <- as.matrix(ref_sc@assays$SCT@counts) # extracts counts matrix of reference data set
ref_meta <- as.data.frame(ref_sc@meta.data) # extracts meta_data (barcodes, clusters, and nUMI)

cell_types <- ref_meta$seurat_clusters
names(cell_types) <- rownames(ref_meta) # create cell_types named list
cell_types <- as.factor(cell_types) # convert to factor data type

nUMI <- colSums(ref_counts); 
names(nUMI) <- rownames(ref_meta) # create nUMI named list

### Create the Reference object
reference <- Reference(ref_counts, cell_types, nUMI)
blocs<- c("bloc2", "bloc4")

counts <- as.matrix(combined@assays$RNA$counts)# extracts counts matrix of spatial data (before Filtering)

rm(combined,ref_sc, ref_counts)
gc()

RCTD_list<-list()
for (i in 1:length(blocs)){
   counts_filt<- counts[,grep(pattern=blocs[i], colnames(counts))]
   coords_filt<- coords[grep(pattern=blocs[i], rownames(coords)),]
   nUMI_tmp <- colSums(counts) # In this case, total counts per pixel is nUMI

   ### Create SpatialRNA object
   puck <- SpatialRNA(coords_filt, counts_filt, nUMI_tmp)
   RCTD_list[[i]]<- create.RCTD(puck, reference, max_cores = 4,UMI_min = min(nUMI_tmp))
   RCTD_list[[i]]<- run.RCTD( RCTD_list[[i]], doublet_mode = 'full')
 }

```

```{r RCTD_load}
RCTD<-readRDS(file.path(projDir,"/Data/Spatial/RCTD_perbloc.rds"))
```

```{r RCTD_Assign}
### extract cell type assignments
RCTD_assign_list<-list()
for (i in 1:3){
  RCTD_assign_list[[i]]<-ctype_assign_RCTD(RCTD[[i]])
}
RCTD_assign_df<- bind_rows(RCTD_assign_list)

### assign spots with several assignments as "Multiple"
RCTD_assign_df$cell_types_plot[RCTD_assign_df$Nr_Celltypes == 1]<-paste0("An",RCTD_assign_df$cell_types[RCTD_assign_df$Nr_Celltypes == 1])
RCTD_assign_df$cell_types_plot[RCTD_assign_df$Nr_Celltypes > 1]<-"Multiple"
colnames(RCTD_assign_df)<-paste0(colnames(RCTD_assign_df), "_RCTD")

### Add assignemnt to meta data of seurat object
combined_filt<- AddMetaData(combined_filt, RCTD_assign_df)

### assign spots with no assignments as "Not_Assigned"
combined_filt@meta.data$Nr_Celltypes_RCTD[is.na(combined_filt@meta.data$Nr_Celltypes_RCTD)]<-0
combined_filt@meta.data$cell_types_RCTD[is.na(combined_filt@meta.data$cell_types_RCTD)]<-"Not_Assigned"
combined_filt@meta.data$cell_types_plot_RCTD[is.na(combined_filt@meta.data$cell_types_plot_RCTD)]<-"Not_Assigned"

### relevel factors
combined_filt@meta.data$cell_types_plot_RCTD<- factor(combined_filt@meta.data$cell_types_plot_RCTD, levels=c("An0","An1","An2","An3","An4","An5","An6","An7","An8","An9","An10","An11","An12","An13","An14","An15","An16","An17","An18","An19","An20","An21","An22","An23","An24","An25","An26","An27","An28","An29","An30","An31","An32","An33","An34","An35","An36","An37","An38","An39","An40","An41","An42","Multiple","Not_Assigned"))

```


# Step 7: Normalization and Clustering
```{r Norm_Clust}
### Noramlisation with scTransfrorm
combined_filt <- SCTransform(combined_filt, vars.to.regress = c("percent.mt", "bloc"), verbose = FALSE )
### Run PCA
combined_filt <- RunPCA(combined_filt, verbose = FALSE, seed.use = 42)
### Run UMAP
combined_filt <- RunUMAP(combined_filt, dims = 1:40, verbose = FALSE, seed.use = 1866,reduction.name ="umap")
### Clustering
combined_filt <- FindNeighbors(combined_filt, dims = 1:50, verbose = FALSE )
combined_filt <- FindClusters(combined_filt, verbose = FALSE, resolution = 0.5,random.seed = 0, cluster.name = "Sp_clusters")
```

```{r dimred_addmeta}
umapdf<- as.data.frame(combined_filt@reductions$umap@cell.embeddings)
pcadf<- as.data.frame(combined_filt@reductions$pca@cell.embeddings[,1:3])


combined_filt<- AddMetaData(combined_filt, pcadf)
combined_filt<- AddMetaData(combined_filt, umapdf)
```

# Step 8: Subclustering of cluster Sp2
```{r  Subset_Sp2}
counts_Sp2<- counts[,colnames(counts) %in% combined_filt@meta.data$XC[combined_filt@meta.data$Sp_clusters ==2]]
counts_Sp2<- counts_Sp2[rowSums(counts_Sp2) > 0,]
combined_filt_Sp2<-CreateSeuratObject(counts=counts_Sp2, meta.data = combined_filt@meta.data[combined_filt@meta.data$Sp_clusters ==2,], project = "SpatialTranscriptomics_Sp2")
```


```{r Norm_Clust_Sp2}
combined_filt_Sp2 <- SCTransform(combined_filt_Sp2, vars.to.regress = c("bloc", "percent.mt"), verbose = FALSE)
combined_filt_Sp2 <- RunPCA(combined_filt_Sp2, verbose = FALSE)
combined_filt_Sp2 <- RunUMAP(combined_filt_Sp2, dims = 1:50, verbose = FALSE, seed.use = 1866)
combined_filt_Sp2 <- FindNeighbors(combined_filt_Sp2, dims = 1:50, verbose = FALSE)
combined_filt_Sp2 <- FindClusters(combined_filt_Sp2, verbose = FALSE, resolution = 1.2,random.seed = 1866)
```

# Step 9: Assignment of Kenyan cells, MIP cells and octopaminergic cells in cluster Sp2
```{r KC_Assign}
mushypees<-combined_filt@meta.data$XC[combined_filt@meta.data$umap_1 < 2.5 & combined_filt@meta.data$umap_2 > 7.75]
combined_filt_Sp2@meta.data$Mushroom<- "No"
combined_filt_Sp2@meta.data$Mushroom[combined_filt_Sp2@meta.data$XC %in% mushypees] <-"Yes"

combined_filt@meta.data$Mushroom<- "No"
combined_filt@meta.data$Mushroom[combined_filt@meta.data$XC %in% mushypees] <-"Yes"
```

```{r MIP_Assign}
MIpees<-combined_filt@meta.data$XC[combined_filt@meta.data$umap_1 < 1.8 & combined_filt@meta.data$umap_2 > 6.5]

combined_filt_Sp2@meta.data$MIPcell<- "No"
combined_filt_Sp2@meta.data$MIPcell[combined_filt_Sp2@meta.data$XC %in% MIpees] <-"Yes"

combined_filt@meta.data$MIPcell<- "No"
combined_filt@meta.data$MIPcell[combined_filt@meta.data$XC %in% MIpees] <-"Yes"
```

```{r Octo_Assign}
Octopees<-combined_filt@meta.data$XC[combined_filt@meta.data$umap_1 > 4.9 & combined_filt@meta.data$umap_2 > 4]

combined_filt_Sp2@meta.data$Octopaminergic<- "No"
combined_filt_Sp2@meta.data$Octopaminergic[combined_filt_Sp2@meta.data$XC %in% Octopees] <-"Yes"

combined_filt@meta.data$Octopaminergic<- "No"
combined_filt@meta.data$Octopaminergic[combined_filt@meta.data$XC %in% Octopees] <-"Yes"
```

# Step 10: Differential gene expression analysis
```{r DEG_Sp, eval=FALSE}
DEGs_SpClusters<- GetAllMarkers(combined_filt, group = "Sp_clusters" , method = "MAST")
```

```{r DEG_KC, eval=FALSE}
p.th<- 0.05
lf.th<-0.5
pct.th<- 0.25
DEGs_Mushroom<-list()
DEGs_Mushroom$AllTests<-FindMarkers(combined_filt_Sp2, ident.1 = "Yes", ident.2 = "No", group.by = "Mushroom", min.pct = 0, min.cells.feature = 0, min.cells.group = 0, recorrect_umi=FALSE, logfc.threshold = 0, test.use="MAST")

DEGs_Mushroom$AllTests$Gene.stable.ID<- rownames(DEGs_Mushroom$AllTests)

DEGs_Mushroom$AllTests<- dplyr::left_join(DEGs_Mushroom$AllTests, All_genes_AgamP4_PEST, by="Gene.stable.ID")
DEGs_Mushroom$AllTests$Gene.name[is.na(DEGs_Mushroom$AllTests$Gene.name)]<- DEGs_Mushroom$AllTests$Gene.stable.ID[is.na(DEGs_Mushroom$AllTests$Gene.name)]

DEGs_Mushroom$Significant<-DEGs_Mushroom$AllTests[DEGs_Mushroom$AllTests$p_val_adj < p.th & abs(DEGs_Mushroom$AllTests$avg_log2FC) > lf.th, ]
DEGs_Mushroom$Filtered<-DEGs_Mushroom$Significant[DEGs_Mushroom$Significant$pct.1 > pct.th | DEGs_Mushroom$Significant$pct.2 > pct.th  , ]
```

```{r DEG_MIP, eval=FALSE }
p.th<- 0.05
lf.th<-0.5
pct.th<- 0.25

DEGs_MIP<-list()
DEGs_MIP$AllTests<-FindMarkers(combined_filt_Sp2, ident.1 = "Yes", ident.2 = "No", group.by = "MIPcell", min.pct = 0, min.cells.feature = 0, min.cells.group = 0, recorrect_umi=FALSE, logfc.threshold = 0, test.use="MAST")

DEGs_MIP$AllTests$Gene.stable.ID<- rownames(DEGs_MIP$AllTests)

DEGs_MIP$AllTests<- dplyr::left_join(DEGs_MIP$AllTests, All_genes_AgamP4_PEST, by="Gene.stable.ID")
DEGs_MIP$AllTests$Gene.name[is.na(DEGs_MIP$AllTests$Gene.name)]<- DEGs_MIP$AllTests$Gene.stable.ID[is.na(DEGs_MIP$AllTests$Gene.name)]

DEGs_MIP$Significant<-DEGs_MIP$AllTests[DEGs_MIP$AllTests$p_val_adj < p.th & abs(DEGs_MIP$AllTests$avg_log2FC) > lf.th, ]
DEGs_MIP$Filtered<-DEGs_MIP$Significant[DEGs_MIP$Significant$pct.1 > pct.th | DEGs_MIP$Significant$pct.2 > pct.th  , ]
```

```{r DEG_Octo, eval=FALSE}
p.th<- 0.05
lf.th<-0.5
pct.th<- 0.25

DEGs_Octo<-list()
DEGs_Octo$AllTests<-FindMarkers(combined_filt_Sp2, ident.1 = "Yes", ident.2 = "No", group.by = "Octopaminergic", min.pct = 0, min.cells.feature = 0, min.cells.group = 0, recorrect_umi=FALSE, logfc.threshold = 0, test.use="MAST")

DEGs_Octo$AllTests$Gene.stable.ID<- rownames(DEGs_Octo$AllTests)

DEGs_Octo$AllTests<- dplyr::left_join(DEGs_Octo$AllTests, All_genes_AgamP4_PEST, by="Gene.stable.ID")
DEGs_Octo$AllTests$Gene.name[is.na(DEGs_Octo$AllTests$Gene.name)]<- DEGs_Octo$AllTests$Gene.stable.ID[is.na(DEGs_Octo$AllTests$Gene.name)]

DEGs_Octo$Significant<-DEGs_Octo$AllTests[DEGs_Octo$AllTests$p_val_adj < p.th & abs(DEGs_Octo$AllTests$avg_log2FC) > lf.th, ]
DEGs_Octo$Filtered<-DEGs_Octo$Significant[DEGs_Octo$Significant$pct.1 > pct.th | DEGs_Octo$Significant$pct.2 > pct.th  , ]
```

# Step 11: Spatial Cluster Distances
```{r Plots_load}
combined_filt<-readRDS(file.path(projDir, "Data/Spatial/combined_filtered_seurat.rds"))
```

```{r Clust_Dist}
dist_list<-list()
distances_df<-data.frame()
HQ_Brains<-  HQ_df$Brain_ID[HQ_df$HQ_Brain == "Yes"]
HQ_cluster<- unique(combined_filt@meta.data$Sp_clusters[combined_filt@meta.data$HQ_Brain =="Yes"]) %>% discard(is.na) %>% sort()

for (b in HQ_Brains){
  dist_list[[b]]<-list()
  info_filt_tmp<-filter(combined_filt@meta.data, Brain_ID == b)
  for (i in 1:length(HQ_cluster)){
    dist_list[[b]][[i]]<-list()
    tmp_1<-info_filt_tmp[info_filt_tmp$Sp_clusters == i-1, c("XC","x", "y")]
    for (j in 1:length(HQ_cluster)){
      tmp_2<-info_filt_tmp[info_filt_tmp$Sp_clusters == j-1, c("XC","x", "y")]
      dist_list[[b]][[i]][[j]]<-fields::rdist(tmp_1[,c("x", "y")], tmp_2[,c("x", "y")])
      colnames(dist_list[[b]][[i]][[j]])<-tmp_2$XC
      rownames(dist_list[[b]][[i]][[j]])<-tmp_1$XC

      if (i==j){
        if (ncol(dist_list[[b]][[i]][[j]])>1){
          tmp_df<-data.frame(Brain_ID=b,Cluster_comp1= i-1, Cluster_comp2= j-1, mean_min=mean(Rfast::rownth(dist_list[[b]][[i]][[j]],elems=rep(2,nrow(dist_list[[b]][[i]][[j]])))))  
        }else{
          tmp_df<-data.frame(Brain_ID=b,Cluster_comp1= i-1, Cluster_comp2= j-1, mean_min=NA) 
        }
        
      }else{
        tmp_df<-data.frame(Brain_ID=b,Cluster_comp1= i-1, Cluster_comp2= j-1, mean_min=mean(rowMins(dist_list[[b]][[i]][[j]]))) 
      }
      distances_df<-rbind(distances_df, tmp_df)
    }
  } 
}
distances_df<-distances_df[!(is.na(distances_df$Brain_ID)),]
distances_df<-distances_df[distances_df$mean_min !=Inf,]

distances_df<-filter(distances_df, Cluster_comp1 != 11 & Cluster_comp2 != 11)

distances_df_mean<- distances_df %>% group_by(Cluster_comp1, Cluster_comp2) %>% summarise(mean_min= mean(mean_min)) %>% as.data.frame()
distances_df_mean <- distances_df_mean %>%   rowwise() %>%   mutate(Comparison = paste(sort(c_across(starts_with("Cluster_"))),collapse = "_"))
distances_df_mean2 <- distances_df_mean %>% group_by(Comparison) %>% summarise(mean_min_Comparison= mean(mean_min)) %>% as.data.frame()
distances_df_mean2$Comp1<-as.numeric(sapply(distances_df_mean2$Comparison, function(x){strsplit(x, split = "_")[[1]][1]}))
distances_df_mean2$Comp2<-as.numeric(sapply(distances_df_mean2$Comparison, function(x){strsplit(x, split = "_")[[1]][2]}))

distances_df_mean2$Comp1<-factor(as.character(distances_df_mean2$Comp1), levels=c("0","1","2","3","4","5","6","7","8","9", "10","12"))
distances_df_mean2$Comp2<-factor(as.character(distances_df_mean2$Comp2), levels=c("0","1","2","3","4","5","6","7","8","9", "10","12"))

```

# Step 12: Spatial Expression within the eye
```{r Eye}
zoom_df<- data.frame(Brain_ID=c ("bloc4_4"),
                      side=c("right"),
                      limit_x_rot=c(-75),
                      limit_y_rot=c( 100),
                      center_x_rot=c(-75),
                      center_y_rot=c(100))

# Create custm gene list starting from providing gene names, requires gene annotation
customgenepanel<-All_genes_AgamP4_PEST$Gene.stable.ID[All_genes_AgamP4_PEST$Gene.name %in% c( "CPR132",  "ARR2", "GPROP8","AGAP002291", "AGAP001913", "AGAP000965")]

# Run function to get the expression by distance and get the plots
custom_EOTM<-EotM(object = combined_filt, cluster = c(0,1,2,3,4,5,6,7,8,9,10,11,12), paramdf = zoom_df, genes = customgenepanel, gorder= c( "AGAP000965","AGAP002291","AGAP001913","ARR2", "CPR132", "GPROP8"))

# Model expression of all genes within sections
Fit_all<-EotM_fit(object = combined_filt,cluster = c(0,1,2,3,4,5,6,7,8,9,10,11,12), paramdf = zoom_df, genes = rownames(combined_filt@assays$SCT$data))
```

# Step 13: Plots
## Clustering {.tabset .tabset-dropdown}
### Principal Component Analysis {.tabset}
```{r Plot_PCA, fig.width=8, fig.height=6}
DimPlot(combined_filt, ,label = TRUE, reduction = "pca",group.by = "Sp_clusters", cols = cols_cl)
```

### UMAP {.tabset}
```{r Plot_UMAP, fig.width=8, fig.height=6}
DimPlot(combined_filt, ,label = TRUE, reduction = "umap",group.by = "Sp_clusters", cols = cols_cl)
```

### Spatial {.tabset}
```{r Plot_clSpatial, fig.width=10, fig.height=4}

a<-ggplot()+
        geom_point(data=filter(combined_filt@meta.data, bloc =="bloc2"), aes(x=y, y=x, colour = Sp_clusters), size=0.1)+
        theme_bw()+
        scale_y_reverse()+
        scale_color_manual(values = cols_cl, name="Sp Cluster")+
        guides(colour = FALSE)+
        ylab("")+xlab("")
b<-ggplot()+
        geom_point(data=filter(combined_filt@meta.data, bloc =="bloc2"), aes(x=y, y=x, colour = Sp_clusters), size=0.1)+
        theme_bw()+
        scale_y_reverse()+
        scale_color_manual(values = cols_cl, name="Sp Cluster")+
        guides(colour = guide_legend(override.aes = list(size=5), ncol=2))+
        ylab("")+xlab("")

cowplot::plot_grid(a,b, rel_widths = c(0.8,1))
```

### Single Brains {.tabset}
```{r Plot_Clbrains, fig.width=10, fig.height=8}
plot_list<-list()
for (i in Brains){
  df_plot<-filter(combined_filt@meta.data, Brain_ID == i)
  plot_list[[i]]<-ggplot()+
                    geom_point(data=df_plot, aes(x=y_rot, y=x_rot, colour = Sp_clusters), size=0.5)+
                    theme_bw()+
                    scale_y_reverse()+
                    scale_color_manual(values = cols_cl, name="Sp Cluster")+
                    guides(colour = FALSE)+
                    coord_fixed()+
                    ylab("")+xlab("")
}
legend <- cowplot::get_legend(b)
cowplot::plot_grid(legend, plotlist = plot_list)
                 
```

## Deconvolution {.tabset}
```{r Plots_Decon, fig.height=10, fig.width=15}
df<- as.data.frame(table(combined_filt@meta.data$Sp_clusters, combined_filt@meta.data$cell_types_plot_RCTD))

colnames(df)<-c("Sp_clusters", "RCTD", "Nr_spots")
df$RCTD<- factor(df$RCTD, levels=c("An0","An1","An2","An3","An4","An5","An6","An7","An8","An9","An10","An11","An12","An13","An14","An15","An16","An17","An18","An19","An20","An21","An22","An23","An24","An25","An26","An27","An28","An29","An30","An31","An32","An33","An34","An35","An36","An37","An38","An39","An40","An41","An42","Multiple","Not_Assigned"))



ggplot()+
  geom_alluvium(data=df,aes(y = Nr_spots, axis1 = Sp_clusters, axis2 = RCTD, fill = Sp_clusters), width = 1/12) +
  geom_stratum(data=df,aes(y = Nr_spots, axis1 = Sp_clusters, axis2 = RCTD),width = 1/12,alpha = .5) +
  geom_label(data=df, aes(y = Nr_spots, axis1 = Sp_clusters, axis2 = RCTD,label = after_stat(stratum)),stat = "stratum") +
  scale_x_discrete(limits = c("Sp_clusters", "RCTD"), expand = c(.05, .05)) +
  scale_fill_manual(values=cols_cl) +
    theme_bw()
```

## Cluster Distance {.tabset}
```{r Plots_Clust_Dif, fig.height=7, fig.width=7}
ggplot()+
  geom_tile(data=distances_df_mean2, aes(x=Comp1, y=Comp2, fill=mean_min_Comparison*0.647))+
  scale_fill_gradientn(colours = rev(c("blue4","blue","lightblue","orange", "red","red4")), trans="log2", name = "Mean Minimal Distance in µm")+
  theme_bw()+
  coord_fixed()+
  xlab("Spatial Cluster")+ylab("Spatial Cluster")
```

## Eye section {.tabset}
### Example of Spatial Markers {.tabset}
```{r Plots_SpatialMarkers, fig.height=7, fig.width=15}
a<-cowplot::plot_grid(custom_EOTM$right$bloc4_4$plots$Selection, custom_EOTM$right$bloc4_4$plots$DistanceEdge)
cowplot::plot_grid(a, custom_EOTM$right$bloc4_4$plots$Geneplot_spatial, custom_EOTM$right$bloc4_4$plots$Distancebox_edge,custom_EOTM$right$bloc4_4$plots$Geneplot_edge, ncol=2)
```

### Spatially varbiable genes  {.tabset}
```{r Plots_SpatialVariable, fig.height=10, fig.width=5}
filt_expr<-list()
for (i in 1:length(Fit_all$right)){
  data_long<-filter(Fit_all$right[[i]]$data$ExpressionData, Gene.stable.ID %in% Fit_all$right[[i]]$data$NatSpline_stats$.id[Fit_all$right[[i]]$data$NatSpline_stats$adj.r_squ > 0.1])
  expr_ranges_filt<-filter(Fit_all$right[[i]]$data$expr_ranges, Max-Min >1.5)
  filt_expr[[i]]<-filter(data_long, Gene.stable.ID %in% expr_ranges_filt$Gene.stable.ID)
  max_dist<- max(filt_expr[[i]]$distance_0)
  filt_expr[[i]]$distance_from_edge<- max_dist-filt_expr[[i]]$distance_0
  names(filt_expr)[i]<- names(Fit_all$right)[i]
}

filt_expr_comb<-bind_rows(filt_expr, .id = "Brain_ID")

Brain_genes<-unique(filt_expr_comb[, c("Brain_ID", "Gene.name")])
Brain_genes<- as.data.frame(table(Brain_genes$Gene.name))

### model the clusters distribution
dist_clus_filt<-list()
for (i in 1:length(Fit_all$right)){
  zoom<- Fit_all$right[[i]]$data$zoom
  zoom$Distance_bin<-cut(x = zoom$distance_0,breaks=seq(1,max(zoom$distance_0)+10, 10))
  dist_clus<-as.data.frame(table(zoom$Distance_bin, zoom$Sp_clusters))
  colnames(dist_clus)<-c("Distance_bin",  "Sp_clusters",  "Freq")
  dist_clus$Distance_bin_numeric<-sapply(as.character(dist_clus$Distance_bin), function(x){as.numeric(strsplit(strsplit(x, split = ",", fixed = T)[[1]][1], split = "(", fixed = T)[[1]][2])})
  dist_clus_filt[[i]]<-filter(dist_clus,Sp_clusters %in% c(0,1,2,4,5,8) )
  max_dist<- max(dist_clus_filt[[i]]$Distance_bin_numeric)
  dist_clus_filt[[i]]$distance_from_edge<- max_dist-dist_clus_filt[[i]]$Distance_bin_numeric
  names(dist_clus_filt)[i]<- names(Fit_all$right)[i]
  }

dist_clus_filt_com<-bind_rows(dist_clus_filt, .id = "Brain_ID")

dist_clus_filt_com_mean<- dist_clus_filt_com %>% group_by(Sp_clusters, distance_from_edge) %>% summarise(Mean_Freq=mean(Freq)) %>% as.data.frame()

### Get the peak from the modeled expression
assigned_list<-list()
for (i in 1:length(Fit_all$right)){
  assignedDF<-data.frame()
  for (j in unique(filt_expr[[i]]$Gene.stable.ID)){
    daje<-data.frame(distance_0=seq(from=1, to= max(Fit_all$right[[i]]$data$zoom$distance_0), by=10),norm_Expr=predict(Fit_all$right[[i]]$NatSplineFits[[j]], newdata=data.frame(distance_0=seq(from=1, to=max(Fit_all$right[[i]]$data$zoom$distance_0),, by=10))))
    peaks<- daje %>% top_n(wt = norm_Expr,n = 1) %>% .$distance_0
    
    assigned_clusters<-filter(dist_clus_filt[[i]], Distance_bin_numeric == peaks ) %>% top_n(wt = Freq, n =1 ) %>% .$Sp_clusters
    max_dist<- max(Fit_all$right[[i]]$data$zoom$distance_0)
    ascl<-paste0(assigned_clusters, collapse = "_")
    ascl_peaks<-peaks
    ascl_peaksedge<-max_dist-peaks
    tmpdf<- data.frame(assigned_clusters=ascl,Gene.stable.ID=j, peak_bin=ascl_peaks, peak_bin_from_edge= ascl_peaksedge)
    
    assignedDF<-rbind(assignedDF, tmpdf)
  }
  assigned_list[[i]]<-dplyr::left_join(filt_expr[[i]], assignedDF, by="Gene.stable.ID")
  names(assigned_list)[i]<- names(Fit_all$right)[i]
}

assigned_list_com<-bind_rows(assigned_list, .id = "Brain_ID")


assigned_list_mean <- assigned_list_com[, c("Gene.name","Gene.stable.ID", "peak_bin_from_edge", "norm_Expr")] %>% group_by(Gene.name,Gene.stable.ID) %>% summarise(mean_peak_bin_from_edge= mean(peak_bin_from_edge), mean_norm_Expr= mean(norm_Expr))
assigned_list_com_wide <- unique(assigned_list_com[, c("Gene.name", "Brain_ID", "assigned_clusters")]) %>% pivot_wider(names_from = Brain_ID, values_from =assigned_clusters )

assigned_list_mean_Brain_genes<- filter(assigned_list_mean, Gene.name %in% Brain_genes$Var1[Brain_genes$Freq > 1])

ggplot()+
  geom_smooth(data = dist_clus_filt_com_mean, aes(x=distance_from_edge*0.647, y=Mean_Freq, colour=Sp_clusters),method = "lm",formula = y ~ ns(x, 10), se=F)+
  geom_label_repel(data = assigned_list_mean_Brain_genes, aes(x=mean_peak_bin_from_edge*0.647, y=3*mean_norm_Expr, label=Gene.name),size=3, max.overlaps = 500)+
  theme_bw()+  scale_color_manual(values=cols_cl, name=("Spatial Clusters"))+ scale_x_continuous(breaks = c(0,50,100,150,200,250,300))+
  scale_y_continuous(name = "Cluster Abundance", sec.axis = sec_axis(~./3, name="Mean normalized Expression")  )+xlab("Mean distance from edge of the eye [µm]")+coord_flip()+
  guides(colour = guide_legend(override.aes = list(size=10)))
```

# Session Info
```{r sessioninfo}
sessionInfo()
```




