---
title: 'The Anopheles mosquito brain at spatial and single-cell resolution: \n Analyis of Single Cell Transcriptomics Data'
author: "Irene Rossi, Johannes W. Bagnoli"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    number_sections: false
    keep_md: false
    self_contained: yes
---

# Step 1: load required packages & functions:
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = F, error = F)
```

```{r seed}
set.seed(666)
projDir<-here::here()
```

```{r packages, results='hide'}
libs <- c("Seurat","dplyr","SeuratData", "SeuratWrappers", "limma","scCustomize","patchwork","ggplot2", "cowplot","glmGamPoi", "DoubletFinder", "ROCR", "Matrix", "plyr", "pagoda2", "tidyverse", "MAST", "grid", "viridis", "scales", "reshape2")
lapply(libs, library, character.only = TRUE)
```

```{r functions_preprocess}
preprocess<- function(paramdf){
  object_list<-list()
  for (i in 1:nrow(paramdf)){
    print(paramdf$file[i])
    ## Creating Seurat objects, add meta data, and filter
    object_list[[i]] <- CreateSeuratObject(counts = Read_CellBender_h5_Mat(paramdf$file[i]), min.cells = paramdf$min.cells[i], min.features = paramdf$min.features[i])
    object_list[[i]] <- RenameCells(object = object_list[[i]], add.cell.id = paramdf$sample[i])
    object_list[[i]]@meta.data$sample <- paramdf$sample[i]
    object_list[[i]]@meta.data$condition <- paramdf$condition[i]
    object_list[[i]]@meta.data$percent_apopt <- PercentageFeatureSet(object_list[[i]], features = intersect(mt_vector, rownames(object_list[[i]])))
    object_list[[i]] <- subset(object_list[[i]], nFeature_RNA > paramdf$nFeature[i] & nCount_RNA > paramdf$nCount[i] & percent_apopt < 20)
    ## sctransform
    object_list[[i]] <- SCTransform(object_list[[i]], vst.flavor = "v2", vars.to.regress = "percent_apopt", verbose = FALSE) # you can ignore the warnings
    object_list[[i]] <- RunPCA(object_list[[i]], npcs = 40, verbose = FALSE)
    object_list[[i]] <- RunUMAP(object_list[[i]], reduction = "pca", dims = 1:40, verbose = FALSE)
    object_list[[i]] <- FindNeighbors(object_list[[i]], reduction = "pca", dims = 1:40, verbose = FALSE)
    object_list[[i]] <- FindClusters(object_list[[i]], resolution = 1, verbose = FALSE)
    ## Sweep thingy
    sweep.res <- DoubletFinder::paramSweep(object_list[[i]], sct = T)
    sweep.stats <- summarizeSweep(sweep.res, GT = FALSE)
    bcmvn <- find.pK(sweep.stats)
    for_pK <- bcmvn[bcmvn$BCmetric == max(bcmvn$BCmetric), ]
    pK <- as.vector(for_pK$pK[[1]])
    pK <- as.numeric(pK)
    nExp_poi <- round(0.15*nrow(object_list[[i]]@meta.data))
    object_list[[i]] <- DoubletFinder::doubletFinder(object_list[[i]], pK = pK, nExp = nExp_poi, sct = T, PCs = 1:10)
    DF.name <- colnames( object_list[[i]]@meta.data)[grepl("DF.classification", colnames(object_list[[i]]@meta.data))]
    object_list[[i]] <- object_list[[i]][, object_list[[i]]@meta.data[, DF.name] == "Singlet"]
    ## Save filter objects
    saveRDS(object_list[[i]] , paste0(projDir,"/Data/SingleCell/cellbender/filtered/cb_", paramdf$sample[i], "_filtered.rds") )
  }
  ## integrate ----
  mating.cb.features <- SelectIntegrationFeatures(object.list = object_list, nfeatures = 3000)
  object_list <- PrepSCTIntegration(object.list = object_list, anchor.features = mating.cb.features)
  ## find anchors and integrate
  mating.cb.anchors <- FindIntegrationAnchors(object.list = object_list, normalization.method = "SCT", anchor.features = mating.cb.features)
  mating.cb.sct <- IntegrateData(anchorset = mating.cb.anchors, normalization.method = "SCT") # the normalization method used is specified
  return(mating.cb.sct)
}
```

```{r functions_DEGcomp}
# DEA and comparison to bulk RNA-seq significant DEG lists
# Function to run DEG analysis on seurat object per cluster between conditions
## Params
### object: seurat object
### Cons: name of column having condition to be checked against each other
### Uno: name of query condition to be checked
### Due: name of baseline condition to be checked
### subs: name of column harboring the identity in which DEG should be performed, default seurat_clusters
### th.fc: log fold change threshold, default 0.25
### th.pval: p value threshold, default 0.05
### th.pct: percentage of cells expressing the genes threshold, default 0.25
### testo: test to be used, default MAST
### subs_prefix: character of prefix for naming of subs, default "Cluster_"
### do_general: logical, whether to perform DEG analysis also on complete data set without sub-setting, default TRUE
FindMarkers_perCond<- function(object,Cons,Uno,Due,subs,th.fc = 0.25,th.pval = 0.05,th.pct = 0.25,testo = "MAST", subs_prefix="Cluster_", do_general = TRUE){
  Idents(object) <- object@meta.data[[subs]]
  loops<- length(unique(Idents(object)))
  DEG_list<-list()
  for (i in 1:loops){
    print(paste0(subs_prefix, sort(unique(Idents(object)))[i]))
    DEG_list[[paste0(subs_prefix, sort(unique(Idents(object)))[i])]]<-list()
    DEG_list[[paste0(subs_prefix, sort(unique(Idents(object)))[i])]]$Allgenes <- FindMarkers(object=object, ident.1 = Uno, ident.2 = Due, group.by=Cons, subset.ident = sort(unique(Idents(object)))[i], min.pct = 0, min.cells.feature = 0, min.cells.group = 0, recorrect_umi=FALSE, logfc.threshold = 0,test.use=testo)
    DEG_list[[paste0(subs_prefix, sort(unique(Idents(object)))[i])]]$Allgenes$Gene.stable.ID<-rownames(DEG_list[[paste0(subs_prefix, sort(unique(Idents(object)))[i])]]$Allgenes)
    DEG_list[[paste0(subs_prefix, sort(unique(Idents(object)))[i])]]$Significant <- filter(DEG_list[[paste0(subs_prefix, sort(unique(Idents(object)))[i])]]$Allgenes, p_val_adj <= th.pval & abs(avg_log2FC) > th.fc & (pct.1 > th.pct | pct.2 > th.pct ) )
  }
  if (do_general == TRUE){
    print("General DEG")
    DEG_list[["general"]]<-list()
    DEG_list[["general"]]$Allgenes <- FindMarkers(object=object, ident.1 = Uno , ident.2 = Due , group.by=Cons, min.pct = 0, min.cells.feature = 0, min.cells.group = 0, recorrect_umi=FALSE, logfc.threshold = 0, test.use=testo)
    DEG_list[["general"]]$Allgenes$Gene.stable.ID<-rownames(DEG_list[["general"]]$Allgenes)
    DEG_list[["general"]]$Significant <- filter( DEG_list[["general"]]$Allgenes, p_val_adj <= th.pval & abs(avg_log2FC) > th.fc & (pct.1 > th.pct | pct.2 > th.pct ) )
  }
  return(DEG_list)
}
```

```{r functions_labs}
palette <- c("khaki", "#FAEBD7", "#E79EB9", "chocolate4", "darkseagreen", "#D5529C", "#C91C8C", "darkgoldenrod", "burlywood", "#330000", "#FF6600",
                 "lightblue", "turquoise", "#E564D3", "#FF1144", "darkblue", "#D740AF", "turquoise4", "#990066", "#005AB5", "#6699CC", "#F083E4", "#F69CDA",
                 "#FFC20A", "#6BB5FF", "orange", "indianred3", "#E6ACB6", "#D6003D", "#BC8A94", "#66CCFF", "#FCB5CF", "#CC0000", "#936972", "royalblue1",
                 "turquoise3", "#3399FF", "darkgreen", "slateblue4", "#C3CCFF", "blue", "#8B5064", "slateblue", "#8B395B", "#8B2252")
# customize plot with cluster numbers
# order labels numerically
labels <- c(
    "Ag0 Neuroblast",
    "Ag1 Neurons",
    "Ag2 Neurons",
    "Ag3 GABAergic neurons",
    "Ag4 Optic lobe neurons",
    "Ag5 Neurons",
    "Ag6 Neurons",
    "Ag7 Monoaminergic neurons",
    "Ag8 GABAergic neurons",
    "Ag9 Glutamatergic neurons",
    "Ag10 Neuropil-associated glia",
    "Ag11 Lamina neurons",
    "Ag12 Optic lobe neurons",
    "Ag13 Neurons",
    "Ag14 Kenyon cells (type I)",
    "Ag15 Photoreceptor neurons",
    "Ag16 Neurons",
    "Ag17 Optic lobe neurons",
    "Ag18 Neurons",
    "Ag19 Lamina neurons",
    "Ag20 Neurons",
    "Ag21 Neurons",
    "Ag22 Neurons",
    "Ag23 Astrocyte-like glia",
    "Ag24 L4/L5-like neurons",
    "Ag25 Surface glia",
    "Ag26 Glutamatergic and cholinergic neurons",
    "Ag27 Neurons",
    "Ag28 Kenyon cells (type II)",
    "Ag29 Neurons",
    "Ag30 Optic lobe neurons",
    "Ag31 Neurons",
    "Ag32 Kenyon cells (type III)",
    "Ag33 Neurons",
    "Ag34 Optic lobe neurons",
    "Ag35 Optic lobe neurons",
    "Ag36 Optic lobe neurons",
    "Ag37 Clock neurons",
    "Ag38 Oenocytoids",
    "Ag39 Optic lobe neurons (putative C2/C3)",
    "Ag40 Optic lobe neurons (putative Pm)",
    "Ag41 Central complex (poxn neurons)",
    "Ag42 Granulocytes",
    "Ag43 GABAergic neurons",
    "Ag44 GABAergic neurons"
)
# extract the numeric part
numbers <- as.numeric(sub("An(\\d+).*", "\\1", labels))
# sort labels
sorted_labels <- labels[order(numbers)]
```

# Step 2: load data and annotation files
```{r load_mt, eval=FALSE}
## loading mitochondiral gene annotation
mt_gtf <- read.table(file.path(projDir,"/Data/Annotations/AgamP4_Mt.gtf"), header = FALSE, sep = '\t')
df_mt <- as.data.frame(mt_gtf)
mt_vector <- df_mt$V9
mt_vector <- gsub(".*gene_id ", "", as.character(mt_vector))
mt_vector <- gsub(";","", as.character(mt_vector))
mt_vector <- unique(mt_vector)
```

# Step 3: Creation of seurat object, QC and integration of datasets
```{r preprocess, eval=FALSE}
paramdf<-data.frame(file=list.files(paste0(projDir,"/Data/SingleCell/cellbender/"), full.names = T, pattern = "*.h5"),
                    min.cells=c(5,5,5,5,5,5),
                    min.features=c(350,350,350,350,350,350),
                    sample=c("B1", "B2", "M1", "M2","V1", "V2"),
                    condition=c("blood-fed","blood-fed" ,"mated","mated", "virgin","virgin"),
                    nFeature=c(800,800,800,800,800,800),
                    nCount=c(1000,1000,1000,1000,1000,1000)
                    )

mating.cb.sct<-preprocess(paramdf)
```

# Step 4: Normalisation & clustering 
```{r Norm_clust, eval=FALSE}
DefaultAssay(mating.cb.sct) <- "integrated"
mating.cb.sct <- RunPCA(mating.cb.sct, verbose = F, dims = 1:40)
mating.cb.sct <- RunUMAP(mating.cb.sct, seed.use = 123, dims = 1:40)
mating.cb.sct <- FindNeighbors(mating.cb.sct, dims = 1:40)
mating.cb.sct <- FindClusters(mating.cb.sct, random.seed = 123, resolution = 0.7, verbose = F)
```

# Step 5: Find Markers of clusters
```{r Find_Mark, eval=FALSE}
DefaultAssay(mating.cb.sct) <- "SCT"
mating.cb.sct <- PrepSCTFindMarkers(mating.cb.sct, assay = "SCT")

Idents(mating.cb.sct) <- "seurat_clusters"
seu.clust.markers <- FindAllMarkers(mating.cb.sct, only.pos = TRUE,
                                    min.pct = 0.25, logfc.threshold = 0.25, TEST.USE = "MAST")
sig.seu.clust.markers <- seu.clust.markers[seu.clust.markers$p_val_adj<0.05,]
```

## Adding info
```{r ctype_info, eval=FALSE}
Ctype_annotation<-read.csv(file.path(projDir, "/Data/Annotations/best_hits_SAMap.csv"), row.names = 1, sep=",", header=T)

# get matching cluster IDs between Seurat object and Ctype_annotation table
matching_clusters <- intersect(rownames(Ctype_annotation), levels(mating.cb.sct$seurat_clusters))
# add annotation information
for (cluster_id in matching_clusters) {
  cluster_indices <- which(mating.cb.sct$seurat_clusters == cluster_id)
  annotation_info <- Ctype_annotation[cluster_id, ]
  # add annotation information to metadata columns
  mating.cb.sct@meta.data[cluster_indices, "Annotation_cluster" ] <- annotation_info["Annotation_cluster" ]
  mating.cb.sct@meta.data[cluster_indices, "Annotation_cell_type"] <- annotation_info["Annotation_cell_type"]
}
```

# Step 6: Comparison of Differential gene expression between bulk and sinlge cell {.tabset}
```{r DEG_perclusters, eval=FALSE}
virgin <-FindMarkers_perCond(object = mating.cb.sct,
                         Cons ="condition",
                          Uno = "virgin",
                          Due = "mated",
                          subs = "seurat_clusters",
                          th.fc = 0.25,
                          th.pval = 0.05,
                          th.pct = 0.25,
                          testo = "MAST",
                          subs_prefix="Cluster_",
                          do_general = TRUE)

# mated
mated <-FindMarkers_perCond(object = mating.cb.sct,
                               Cons ="condition",
                               Uno = "mated",
                               Due = "virgin",
                               subs = "seurat_clusters",
                               th.fc = 0.25,
                               th.pval = 0.05,
                               th.pct = 0.25,
                               testo = "MAST",
                               subs_prefix="Cluster_",
                               do_general = TRUE)

# blood-fed
bf <-FindMarkers_perCond(object = mating.cb.sct,
                              Cons ="condition",
                              Uno = "blood-fed",
                              Due = "mated",
                              subs = "seurat_clusters",
                              th.fc = 0.25,
                              th.pval = 0.05,
                              th.pct = 0.25,
                              testo = "MAST",
                              subs_prefix="Cluster_",
                              do_general = TRUE)
```

# Step 7: Plotting {.tabset}
```{r Plots_load}
mating.cb.sct.plot<- readRDS(file.path(projDir,"Data/SingleCell/atlas_dataset.rds"))
```

## UMAP {.tabset}
```{r Plots_UMAP}
# reorder the Annotation (cell type) column
mating.cb.sct.plot@meta.data$`Annotation (cell type)` <- factor(
  mating.cb.sct.plot@meta.data$`Annotation (cell type)`,
  levels = sorted_labels
)
# get legend
legend.annotat <- get_legend(DimPlot(mating.cb.sct.plot, cols = palette, reduction = "umap", group.by = "seurat_clusters", label = F))#, na.value = "antiquewhite"))

# get cluster centroids
cluster_centers <- aggregate(Embeddings(mating.cb.sct.plot, "umap"), 
                             by = list(mating.cb.sct.plot@meta.data$seurat_clusters), 
                             FUN = mean)
colnames(cluster_centers) <- c("seurat_clusters", "UMAP_1", "UMAP_2")  # rename columns

Plot_Umap<-DimPlot(mating.cb.sct.plot, cols = palette, reduction = "umap", 
        group.by = "seurat_clusters") +#, na.value = "antiquewhite") + 
  theme(legend.position = "none") +
  geom_text(data = cluster_centers, aes(x = UMAP_1, y = UMAP_2, 
                                        label = seurat_clusters), 
            color = "black", size = 3, fontface = "plain")

cowplot::plot_grid(Plot_Umap, legend.annotat, rel_widths = c(5,1))
```

## Heatmap {.tabset}
```{r Plots_Heatmap, fig.width=20, fig.height=8}
# data to be plotted
data.heatmap <- mating.cb.sct.plot
matrix.data.sct <- data.heatmap@assays$SCT@data
matrix.data.sct <- scale(matrix.data.sct)

data.heatmap@assays$heatmap <- data.heatmap@assays$SCT
data.heatmap@assays$heatmap$scale.data<- matrix.data.sct
DefaultAssay(data.heatmap) <- "heatmap"


new_order <- c(30,24,11:12,4,16:17,19:22,28:29,34,13,18,1:2,5,14,35,32,0,27,6,9,26,36,33,39,3,41,43,44,40,8,7,31,37,15,23,10,25,38,42)
final_colors <- c("#66CCFF","#6BB5FF", "lightblue", "turquoise", "darkseagreen", "#D740AF", "turquoise4",
"#005AB5","#6699CC","#F083E4","#F69CDA","#D6003D","#BC8A94","royalblue1","#E564D3","#990066","#FAEBD7",
"#E79EB9","#D5529C","#FF1144","turquoise3","#CC0000","khaki","#E6ACB6","#C91C8C","#330000","indianred3",
"#3399FF","#936972","#C3CCFF","chocolate4","#8B5064","#8B395B","#8B2252","blue","burlywood","darkgoldenrod",
"#FCB5CF","darkgreen","darkblue","#FFC20A","#FF6600","orange","slateblue4","slateblue")

data.heatmap$seurat_clusters <- factor(data.heatmap$seurat_clusters, levels = new_order)

# glia heatmap of feature expression
# genes and names for all plots will need them
glia.markers <- c("AGAP005639", "AGAP001185", "AGAP001499", "AGAP005223", "AGAP012059", "AGAP002291", "AGAP000278", "AGAP029242", "AGAP011948", "AGAP005563", "AGAP008943")
glia.markers.names <- c("ABC transporter (Family B member 2)", "SCMT1 (AGAP001185)", "AGAP001499", "AGAP005223", "AGAP012059", "ebony", "obp9", "repo", "thr3DH", "tret1",  "turtle")

# then modify the y axis names with cell types in post-production as in cluster_color_mapping_ordered
Plot_heatmap<-DoHeatmap(subset(data.heatmap, downsample = 500), group.colors = final_colors,
          features = glia.markers, size = 3, group.by = "seurat_clusters") + # default slot "scale.data", put slot = data to get just normalized counts
  scale_fill_viridis(option = "B") +
  scale_y_discrete(labels = rev(glia.markers.names)) +
  theme(axis.text.y = element_text(face = "italic")) +
  theme(legend.position="none")
legend.heatmap.glia.markers <- get_legend(DoHeatmap(subset(data.heatmap, downsample = 500), group.colors = final_colors,
                                                    features = glia.markers, size = 3, group.by = "seurat_clusters") +
                                            scale_fill_viridis(option = "B") +
                                            scale_y_discrete(labels = rev(glia.markers.names)) +
                                            theme(axis.text.y = element_text(face = "italic")))
grid.newpage()
cowplot::plot_grid(Plot_heatmap, legend.heatmap.glia.markers, rel_widths = c(10,1))
```

## Differential gene expression {.tabset .tabset-dropdown }
```{r load_DEGloops}
virgin <- readRDS(paste0(projDir,"/Data/SingleCell/DEG/virgin_scMAST.Rds"))
mated <- readRDS(paste0(projDir,"/Data/SingleCell/DEG/mated_scMAST.Rds"))
bf <- readRDS(paste0(projDir,"/Data/SingleCell/DEG/bf_scMAST.Rds"))


# lists of significant DEGs from bulk RNA-seq
DE_MVall <- read.csv(paste0(projDir,"/Data/Bulk/DEG/DE_MVall.csv"), header = T, sep = ",")
DE_BFMall <- read.csv(paste0(projDir,"/Data/Bulk/DEG/DE_BFMall.csv"), header = T, sep = ",")
```

```{r extract_sig}
# upreg in bulk V
DEGbulk_up_V1 <- DE_MVall[DE_MVall$log2FoldChange < 0,]
DEGbulk_up_V <- DEGbulk_up_V1$gene
# upreg in bulk M
DEGbulk_up_M1 <- DE_MVall[DE_MVall$log2FoldChange > 0,]
DEGbulk_up_M <- DEGbulk_up_M1$gene
# upreg in bulk BF
DEGbulk_up_BF1 <- DE_BFMall[DE_BFMall$log2FoldChange > 0,]
DEGbulk_up_BF <- DEGbulk_up_BF1$gene
# downreg in bulk BF
DEGbulk_down_BF1 <- DE_BFMall[DE_BFMall$log2FoldChange < 0,]
DEGbulk_down_BF <- DEGbulk_down_BF1$gene

# Get gene lists ----
# ***either mating-induced (significant)
clust_dea_m <- mated[names(mated) != "general"]

# get significant genes
# if wanting to select for Bonferroni:
m_clust_sig <- lapply(clust_dea_m, function(cluster) {
  if ("Significant" %in% names(cluster)) {
    return(cluster$Significant)
  } else {
    return(NULL)  # or data.frame() for empty clusters
  }
})


# genes m
up_genes_m <- unique(unlist(lapply(m_clust_sig, function(sig_m) {
  if (nrow(sig_m) > 0 && "avg_log2FC" %in% colnames(sig_m)) {
    rownames(sig_m)[sig_m$avg_log2FC > 0]
  } else {
    character(0)
  }
})))

down_genes_m <- unique(unlist(lapply(m_clust_sig, function(sig_m) {
  if (nrow(sig_m) > 0 && "avg_log2FC" %in% colnames(sig_m)) {
    rownames(sig_m)[sig_m$avg_log2FC < 0]
  } else {
    character(0)
  }
})))

# ***or blood meal-induced (significant)
# remove "general" element
clust_dea_bf <- bf[names(bf) != "general"]
# get significant elements
bf_clust_sig <- lapply(clust_dea_bf, function(cluster) {
  if ("Significant" %in% names(cluster)) {
    return(cluster$Significant)
  } else {
    return(NULL)  # or data.frame() for empty clusters
  }
})
# genes bf
up_genes_bf <- unique(unlist(lapply(bf_clust_sig, function(sig_df) {
  if (nrow(sig_df) > 0 && "avg_log2FC" %in% colnames(sig_df)) {
    rownames(sig_df)[sig_df$avg_log2FC > 0]
  } else {
    character(0)
  }
})))

down_genes_bf <- unique(unlist(lapply(bf_clust_sig, function(sig_df) {
  if (nrow(sig_df) > 0 && "avg_log2FC" %in% colnames(sig_df)) {
    rownames(sig_df)[sig_df$avg_log2FC < 0]
  } else {
    character(0)
  }
})))

```

### Dotplots  {.tabset }
#### Mating Status {.tabset }
```{r DEG_Dot_Mating, fig.width=12, fig.height=5}
# mated ----
unique_mated_up <- setdiff(up_genes_m, DEGbulk_up_M)
unique_mated_down <- setdiff(down_genes_m, DEGbulk_up_V)

# Extract data from mated list, including all clusters
cluster_names <- names(m_clust_sig)

# Create a dataframe with counts and percentages
result_df_m <- data.frame()

for (cluster in cluster_names) {
  sig_genes <- m_clust_sig[[cluster]]  # data.frame with rownames
  
  if (nrow(sig_genes) > 0) {
    # Extract each cluster's sig genes
    up_genes <- rownames(sig_genes)[sig_genes$avg_log2FC > 0]
    down_genes <- rownames(sig_genes)[sig_genes$avg_log2FC < 0]
    
    up_count <- length(up_genes)
    down_count <- length(down_genes)
    
    # % overlap with unique genes
    up_pct <- ifelse(up_count > 0, mean(up_genes %in% unique_mated_up) * 100, 0)
    down_pct <- ifelse(down_count > 0, mean(down_genes %in% unique_mated_down) * 100, 0)
    
  } else {
    up_count <- down_count <- 0
    up_pct <- down_pct <- 0
  }
  
  # Add to result dataframe
  temp_df <- data.frame(
    cluster = gsub("Cluster_", "", cluster),
    regulation = c("up", "down"),
    count = c(up_count, down_count),
    pct_unique = c(up_pct, down_pct)
  )
  
  result_df_m <- rbind(result_df_m, temp_df)
}


# Convert cluster to factor for proper ordering + remove general col
cluster_numeric <- as.numeric(gsub(".*?([0-9]+).*", "\\1", result_df_m$cluster))
# general order
order <- c(30,24,11:12,4,16:17,19:22,28:29,34,13,18,
           1:2,5,14,35,32,0,27,6,9,26,36,33,39,3,
           41,43,44,40,8,7,31,37,15,23,10,25,38,42)
result_df_m$cluster <- factor(result_df_m$cluster,
                              levels = order)

# Create the dotplot - size = count, color = % common (viridis turbo)
ggplot(result_df_m, aes(x = cluster, y = regulation, size = count, color = pct_unique)) +
  geom_point() +
  scale_size_continuous(range = c(0, 10), 
                        breaks = c(0, 50, 100, 150, 200),
                        limits = c(0, 200)) +
  scale_color_gradient(low = "aquamarine3", high = "navyblue",
                       name = "% Unique genes") +
  scale_x_discrete(labels = function(x) paste0("Ag", x)) +
  labs(
    x = "Cluster",
    y = "Regulation", 
    size = "Number of genes",
    title = "Significant genes per cluster (% unique genes)",
    color = "% Unique"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.minor = element_blank()
  )
```

#### Feeding Status {.tabset }
```{r DEG_Dot_Feeding, fig.width=12, fig.height=5}
# bf ----
unique_bf_up <- setdiff(up_genes_bf, DEGbulk_up_BF)
unique_bf_down <- setdiff(down_genes_bf, DEGbulk_down_BF)

# Extract data from bf list, including all clusters
cluster_names <- names(clust_dea_bf)

# Create a dataframe with counts and percentages
result_df_bf <- data.frame()

for (cluster in cluster_names) {
  sig_genes <- bf_clust_sig[[cluster]]  # data.frame with rownames
  
  if (nrow(sig_genes) > 0) {
    # Extract each cluster's sig genes
    up_genes <- rownames(sig_genes)[sig_genes$avg_log2FC > 0]
    down_genes <- rownames(sig_genes)[sig_genes$avg_log2FC < 0]
    
    up_count <- length(up_genes)
    down_count <- length(down_genes)
    
    # % overlap with unique genes
    up_pct <- ifelse(up_count > 0, mean(up_genes %in% unique_bf_up) * 100, 0)
    down_pct <- ifelse(down_count > 0, mean(down_genes %in% unique_bf_down) * 100, 0)
    
  } else {
    up_count <- down_count <- 0
    up_pct <- down_pct <- 0
  }
  
  # Add to result dataframe
  temp_df <- data.frame(
    cluster = gsub("Cluster_", "", cluster),
    regulation = c("up", "down"),
    count = c(up_count, down_count),
    pct_unique = c(up_pct, down_pct)
  )
  
  result_df_bf <- rbind(result_df_bf, temp_df)
}

# Convert cluster to factor for proper ordering + remove general col
cluster_numeric <- as.numeric(gsub(".*?([0-9]+).*", "\\1", result_df_bf$cluster))
result_df_bf$cluster <- factor(result_df_bf$cluster,
                               levels = order)


# Create the dotplot - size = count, color = % common (viridis turbo)
ggplot(result_df_bf, aes(x = cluster, y = regulation, size = count, color = pct_unique)) +
  geom_point() +
  scale_size_continuous(range = c(0, 10), 
                        breaks = c(0, 50, 100, 150, 200),
                        limits = c(0, 200)) +
  scale_color_gradient(low = "aquamarine3", high = "navyblue",
                       name = "% Unique genes") +
  scale_x_discrete(labels = function(x) paste0("Ag", x)) +
  labs(
    x = "Cluster",
    y = "Regulation", 
    size = "Number of genes",
    title = "Significant genes per cluster (% unique genes)",
    color = "% Unique"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.minor = element_blank()
  )
```

### Volcano Plots  {.tabset }
#### Mating Status {.tabset }
```{r volc_mating}
# Prepare gene expression data in BFM and sc clusters
# Differential expression status for colours
DE_MVall_plot <- DE_MVall %>%
  dplyr::mutate(
    diffexpressed = dplyr::case_when(
      log2FoldChange > 1.25 ~ "up-regulated",
      log2FoldChange < -1.25 ~ "down-regulated",
      TRUE ~ "log2FC < 1.25"))

# plot
ggplot(DE_MVall_plot,
       aes(x = log2FoldChange,
           y = -log10(padj),
           color = diffexpressed)) +
  geom_hline(yintercept = -log10(0.05), col = "darkgrey", linetype = "dashed") +
  geom_point() +
  scale_color_manual(
    name   = "Differential expression",
    values = c("down-regulated" = "#00AFBB",
               "log2FC < 1.25" = "moccasin",
               "up-regulated" = "red3")
  ) +
  guides(
    color = guide_legend(override.aes = list(size = 1.5)),
    shape = guide_legend(override.aes = list(size = 1.5))
  ) +
  theme_classic() +
  scale_x_continuous() +
  scale_y_log10() +
  ggtitle("Mating effect")

```

#### Feeding Status {.tabset }
```{r, volc_feeding}
# Prepare gene expression data in BFM and sc clusters
# Differential expression status for colours
DE_BFMall_plot <- DE_BFMall %>%
  dplyr::mutate(
    diffexpressed = dplyr::case_when(
      log2FoldChange > 1.25 ~ "up-regulated",
      log2FoldChange < -1.25 ~ "down-regulated",
      TRUE ~ "log2FC < 1.25"))

ggplot(DE_BFMall_plot,
       aes(x = log2FoldChange,
           y = -log10(padj),
           color = diffexpressed)) +
  geom_hline(yintercept = -log10(0.05), col = "darkgrey", linetype = "dashed") +
  geom_point() +
  scale_color_manual(
    name   = "Differential expression",
    values = c("down-regulated" = "#00AFBB",
               "log2FC < 1.25" = "moccasin",
               "up-regulated" = "red3")
  ) +
  guides(
    color = guide_legend(override.aes = list(size = 1.5)),
    shape = guide_legend(override.aes = list(size = 1.5))
  ) +
  theme_classic() +
  scale_x_continuous() +
  scale_y_log10() +
  ggtitle("Blood meal effect")

```


### Heatmap  {.tabset }
```{r Heat_DEG , fig.width=12, fig.height=8}
# gene list
bfm_degs_all <- read.csv(file.path(projDir,"/Data/Bulk/plotDEGs_bulk_bfm.csv"))
gene_list_bf <- bfm_degs_all$gene_id  # your actual gene list in the order you want
# in this case I put DEGs (upreg) in bf

# cluster order
cluster_order <- c(30,24,11:12,4,16:17,19:22,28:29,34,13,18,
                   1:2,5,14,35,32,0,27,6,9,26,36,33,39,3,
                   41,43,44,40,8,7,31,37,15,23,10,25,38,42)

# Create empty dataframe for results
heatmap_data <- data.frame()

# Loop through each cluster
for(cluster_name in names(clust_dea_bf)) {
  # Extract cluster number (removes "Cluster_" prefix)
  cluster_num <- as.numeric(gsub("Cluster_", "", cluster_name))
  
  # Only proceed if cluster is in our order
  if(cluster_num %in% cluster_order) {
    # Get the DE results for this cluster
    de_results <- clust_dea_bf[[cluster_name]]$Allgenes
    
    # For each gene in your list, get the avg_log2FC
    for(gene in gene_list_bf) {
      if(gene %in% rownames(de_results)) {
        fc_value <- de_results[gene, "avg_log2FC"]
      } else {
        fc_value <- NA
      }
      
      # Add to dataframe
      heatmap_data <- rbind(heatmap_data, data.frame(
        Gene = gene,
        Cluster = paste0("Ag", cluster_num),
        avg_log2FC = fc_value,
        stringsAsFactors = FALSE
      ))
    }
  }
}

# Get present clusters
present_clusters <- unique(as.numeric(gsub("Ag", "", heatmap_data$Cluster)))
valid_clusters <- intersect(cluster_order, present_clusters)
cluster_labels <- paste0("Ag", valid_clusters)

# Set factor levels
heatmap_data$Cluster <- factor(heatmap_data$Cluster, levels = cluster_labels)
heatmap_data$Gene <- factor(heatmap_data$Gene, levels = rev(gene_list_bf))

# Create the heatmap with viridis
heatmap_plot_bf <- ggplot(heatmap_data, aes(x = Cluster, y = Gene, fill = avg_log2FC)) +
  geom_tile() +
  scale_fill_viridis(
    option = "B",  # magma
    name = "avg_log2FC",
    na.value = "grey90"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 8),
    axis.text.y = element_text(size = 6),
    axis.title = element_blank(),
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold")
  ) +
  labs(title = "Gene Expression Heatmap (avg_log2FC)")

# Display the plot
print(heatmap_plot_bf)

```



# Session Info
```{r sesion_info}
sessionInfo()
```

