---
title: 'The Anopheles mosquito brain at spatial and single-cell resolution: \n Analysis of infection status and age in Bulk Transcriptomics Data'
author: "Irene Rossi"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    number_sections: false
    keep_md: false
    self_contained: yes
---

# Step 1: load required packages & functions:
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = F, error = F)
```

```{r seed}
set.seed(1234)
projDir<-here::here()
```

```{r packages, results='hide'}

libs <- c("DESeq2","tidyverse","dplyr", "tidyr","reshape2", "readODS", "viridis", "cowplot")
lapply(libs, library, character.only = TRUE)

```

```{r custfunc_PCA13}
alternative.pca1_3 <- function (object, ...) 
{
  .local <- function (object, intgroup = "condition", ntop = 500, 
                      returnData = FALSE) 
  {
    rv <- rowVars(assay(object))
    select <- order(rv, decreasing = TRUE)[seq_len(min(ntop, 
                                                       length(rv)))]
    pca <- prcomp(t(assay(object)[select, ]))
    percentVar <- pca$sdev^2/sum(pca$sdev^2)
    if (!all(intgroup %in% names(colData(object)))) {
      stop("the argument 'intgroup' should specify columns of colData(dds)")
    }
    intgroup.df <- as.data.frame(colData(object)[, intgroup, 
                                                 drop = FALSE])
    group <- if (length(intgroup) > 1) {
      factor(apply(intgroup.df, 1, paste, collapse = ":"))
    }
    else {
      colData(object)[[intgroup]]
    }
    d <- data.frame(PC1 = pca$x[, 1], PC2 = pca$x[, 3], group = group, 
                    intgroup.df, name = colnames(object))
    if (returnData) {
      attr(d, "percentVar") <- percentVar[1:3]
      return(d)
    }
    ggplot(data = d, aes_string(x = "PC1", y = "PC2", color = "group")) + 
      geom_point(size = 3) + xlab(paste0("PC1: ", round(percentVar[1] * 
                                                          100), "% variance")) + ylab(paste0("PC3: ", round(percentVar[3] * 
                                                                                                              100), "% variance")) + coord_fixed() +
      stat_ellipse()
  }
  .local(object, ...)
}
```

# Step 2: load relevant Data and adjust
```{r load_data}
countData <-read.table(file.path(projDir,"/Data/Bulk/Ag_bulk_infections_matrix_corrected.csv"), sep = ",", row.names = 1, header = T)
metaData <- read.table(paste0(projDir,"/Data/Bulk/Ag_infections_phenofile_corrected.csv"), sep = ",", header=T)
```

```{r subset_data}
# create test vs control objects ----
# create an oocyst object
countData_oo <- countData[, grep(pattern = "oocy", colnames(countData))]
metaData_oo <- filter(metaData, sample_name  %in% colnames(countData_oo))

# create replicates objects
countData_oo1 <- countData[, grep(pattern = "oocy1", colnames(countData))]
metaData_oo1 <- filter(metaData, sample_name  %in% colnames(countData_oo1))
countData_oo2 <- countData[, grep(pattern = "oocy2", colnames(countData))]
metaData_oo2 <- filter(metaData, sample_name  %in% colnames(countData_oo2))

# create a sporozoite object
countData_spz <- countData[, grep(pattern = "sporo", colnames(countData))]
metaData_spz <- filter(metaData, sample_name  %in% colnames(countData_spz))

# create replicates objects
countData_spz1 <- countData[, grep(pattern = "sporo1", colnames(countData))]
metaData_spz1 <- filter(metaData, sample_name  %in% colnames(countData_spz1))
countData_spz2 <- countData[, grep(pattern = "sporo2", colnames(countData))]
metaData_spz2 <- filter(metaData, sample_name  %in% colnames(countData_spz2))
```

# Step 3: Analysis using all batches and cell stages
## Using oocyst stage as reference
```{r All_oo}
# make DESeq matrix
dds <- DESeqDataSetFromMatrix(countData = countData, 
                              colData = metaData,
                              design = ~ sample_type, tidy = F)

# set reference level
dds$sample_type <- relevel(dds$sample_type, ref = "control")
# add phenofile cathegories interactions ("design" argument in the above function "DESeqDataSetFromMatrix")
dds$group <- factor(paste0(dds$sample_type, dds$infection_stage))
dds$group <- relevel(dds$group, ref = "controloocyst")
design(dds) <- ~ group
# this way later the result levels would be all contrasted to oocyst controls.

# filter out by at least 3 samples with a count of 10 or higher ----
keep <- rowSums(counts(dds) >= 10) >= 3
dds <- dds[keep,]

# apply DESEQ function ----
dds <- DESeq(dds)

# check oocyst DE
shres_ooall <- lfcShrink(dds, type="apeglm", lfcThreshold=0, 
                         coef=c("group_testoocyst_vs_controloocyst"))

shres_ooall_ashr <- lfcShrink(dds, type="ashr", lfcThreshold=0, 
                              contrast=c("group", "testoocyst", "controloocyst"))

# In nbinomGLM(x = x, Y = YNZ, size = size, weights = weightsNZ, offset = offsetNZ, : the line search routine failed, unable to sufficiently decrease the function value
# usually this rises for extremely low counts - already filtered out - or problematic genes with zero logFC - that were not found in my data.
# Basically it seems that this warning is raised for no reason, so apeglm results are kept for omogeneity with the previous analyses.
# In addition, ashr method is also applied later without raising any error, giving the same DEGs (2 antimicrobial peptides),
# oo (2 replicates) apeglm
sig_shres_ooall <- shres_ooall %>%
  data.frame() %>%
  rownames_to_column(var="gene") %>% 
  as_tibble() %>% 
  filter(padj < 0.05)


# oo (2 replicates) ashr
sig_shres_ooall_ashr <- shres_ooall_ashr %>%
  data.frame() %>%
  rownames_to_column(var="gene") %>% 
  as_tibble() %>% 
  filter(padj < 0.05)


# check sporozoite DE
shres_spz_ashr <- lfcShrink(dds, type="ashr", lfcThreshold=0, 
                            contrast=c("group", "testsporozoite", "controlsporozoite"))
# SPZ
sig_shres_spz_ashr <- shres_spz_ashr %>%
  data.frame() %>%
  rownames_to_column(var="gene") %>% 
  as_tibble() %>% 
  filter(padj < 0.05)


```

## Using sporozoite stage as reference
```{r All_spz}
# see below
dds.spzref <- DESeqDataSetFromMatrix(countData = countData, 
                                     colData = metaData,
                                     design = ~ sample_type, tidy = FALSE)

# set reference level
dds.spzref$sample_type <- relevel(dds.spzref$sample_type, ref = "control")
# add phenofile cathegories interactions ("design" argument in the above function "DESeqDataSetFromMatrix")
dds.spzref$group <- factor(paste0(dds.spzref$sample_type, dds.spzref$infection_stage))
dds.spzref$group <- relevel(dds.spzref$group, ref = "controlsporozoite")
design(dds.spzref) <- ~ group
# this way later the result levels would be all contrasted to sporozoite controls.

# filter out by at least 3 samples with a count of 10 or higher ----
keep <- rowSums(counts(dds.spzref) >= 10) >= 3
dds.spzref <- dds.spzref[keep,]

# apply DESEQ function
dds.spzref <- DESeq(dds.spzref)

# check sporozoite DE
shres_spzall <- lfcShrink(dds.spzref, type="apeglm", lfcThreshold=0, 
                          coef=c("group_testsporozoite_vs_controlsporozoite"))
# spz all
sig_shres_spzall <- shres_spzall %>%
  data.frame() %>%
  rownames_to_column(var="gene") %>% 
  as_tibble() %>% 
  filter(padj < 0.05)
sig_shres_spzall # no DEG
```

# Step 4: Analysis per cell stage
## In oocyst stage
```{r Stage_oo}
# oocyst replicates ----
dds_oo <- DESeqDataSetFromMatrix(countData = countData_oo, 
                                 colData = metaData_oo,
                                 design = ~ sample_type, tidy = FALSE)

# filter out by at least 3 samples with a count of 10 or higher
keep_oo <- rowSums(counts(dds_oo) >= 10) >= 3
dds_oo <- dds_oo[keep_oo,]

# apply DESEQ function
dds_oo <- DESeq(dds_oo)
# shrinked results conditionwise
shres_oo <- lfcShrink(dds_oo, type="apeglm", lfcThreshold=0,
                      coef="sample_type_test_vs_control")

shres_oo_ashr <- lfcShrink(dds_oo, type="ashr", lfcThreshold=0,
                      coef="sample_type_test_vs_control")

# padj < 0.05: subset results to return genes ----
# oo-solo ashr
sig_shres_oo_ashr <- shres_oo_ashr %>%
  data.frame() %>%
  rownames_to_column(var="gene") %>% 
  as_tibble() %>% 
  filter(padj < 0.05)

# oo-solo apeglm
sig_shres_oo <- shres_oo %>%
  data.frame() %>%
  rownames_to_column(var="gene") %>% 
  as_tibble() %>% 
  filter(padj < 0.05)
```

## In sporozoite stage
```{r Stage_spz}
# sporozoite replicates ----
dds_spz <- DESeqDataSetFromMatrix(countData = countData_spz, 
                                  colData = metaData_spz,
                                  design = ~ sample_type, tidy = FALSE)

# filter out by at least 3 samples with a count of 10 or higher
keep_spz <- rowSums(counts(dds_spz) >= 10) >= 3
dds_spz <- dds_spz[keep_spz,]

# apply DESEQ function
dds_spz <- DESeq(dds_spz)

# shrinked results conditionwise
shres_spz_solo <- lfcShrink(dds_spz, type="apeglm", lfcThreshold=0,
                            coef="sample_type_test_vs_control")

shres_spz_solo_ashr <- lfcShrink(dds_spz, type="ashr", lfcThreshold=0,
                            coef="sample_type_test_vs_control")

# padj < 0.05: subset results to return genes ----
# spz-solo ashr
sig_shres_spz_ashr <- shres_spz_solo_ashr %>%
  data.frame() %>%
  rownames_to_column(var="gene") %>% 
  as_tibble() %>% 
  filter(padj < 0.05)

# spz-solo apeglm
sig_shres_spz <- shres_spz_solo %>%
  data.frame() %>%
  rownames_to_column(var="gene") %>% 
  as_tibble() %>% 
  filter(padj < 0.05)
```

# Step 5: Analysis per single batch in both cell stages
## In batch 1 of oocyst stage
```{r batch1_oo}
# * single experimental replicates ----
# batch 1 oo ----
dds_oo1 <- DESeqDataSetFromMatrix(countData = countData_oo1, 
                                  colData = metaData_oo1,
                                  design = ~ sample_type, tidy = FALSE)

# filter out by at least 3 samples with a count of 10 or higher
keep_oo1 <- rowSums(counts(dds_oo1) >= 10) >= 3
dds_oo1 <- dds_oo1[keep_oo1,]

# apply DESEQ function
dds_oo1 <- DESeq(dds_oo1)

# shrinked results conditionwise
shres_oo1 <- lfcShrink(dds_oo1, type="apeglm", lfcThreshold=0,
                       coef="sample_type_test_vs_control")

shres_oo1_ashr <- lfcShrink(dds_oo1, type="ashr", lfcThreshold=0,
                            coef="sample_type_test_vs_control")

# padj < 0.05: subset results to return genes ----
# oo1 ashr
sig_shres_oo1_ashr <- shres_oo1_ashr %>%
  data.frame() %>%
  rownames_to_column(var="gene") %>% 
  as_tibble() %>% 
  filter(padj < 0.05)

# padj < 0.05: subset results to return genes ----
# oo1 apeglm
sig_shres_oo1 <- shres_oo1 %>%
  data.frame() %>%
  rownames_to_column(var="gene") %>% 
  as_tibble() %>% 
  filter(padj < 0.05)
```

## In batch 2 of oocyst stage
```{r batch2_oo}
# batch 2 oo ----
dds_oo2 <- DESeqDataSetFromMatrix(countData = countData_oo2, 
                                  colData = metaData_oo2,
                                  design = ~ sample_type, tidy = FALSE)

# filter out by at least 3 samples with a count of 10 or higher
keep_oo2 <- rowSums(counts(dds_oo2) >= 10) >= 3
dds_oo2 <- dds_oo2[keep_oo2,]

# apply DESEQ function
dds_oo2 <- DESeq(dds_oo2)

# shrinked results conditionwise
shres_oo2 <- lfcShrink(dds_oo2, type="apeglm", lfcThreshold=0, # warning message suggesting possible overparametrization, so check with ashr
                       coef="sample_type_test_vs_control")

shres_oo2_ashr <- lfcShrink(dds_oo2, type="ashr", lfcThreshold=0,
                       coef="sample_type_test_vs_control")

# padj < 0.05: subset results to return genes ----
# oo2 ashr
sig_shres_oo2_ashr <- shres_oo2_ashr %>%
  data.frame() %>%
  rownames_to_column(var="gene") %>% 
  as_tibble() %>% 
  filter(padj < 0.05)

# oo2 apeglm
sig_shres_oo2 <- shres_oo2 %>%
  data.frame() %>%
  rownames_to_column(var="gene") %>% 
  as_tibble() %>% 
  filter(padj < 0.05)
```

## In batch 1 of sporozoite stage
```{r batch1_spz}
# batch 1 spz ----
dds_spz1 <- DESeqDataSetFromMatrix(countData = countData_spz1, 
                                   colData = metaData_spz1,
                                   design = ~ sample_type, tidy = FALSE)

# filter out by at least 3 samples with a count of 10 or higher
keep_spz1 <- rowSums(counts(dds_spz1) >= 10) >= 3
dds_spz1 <- dds_spz1[keep_spz1,]

# apply DESEQ function
dds_spz1 <- DESeq(dds_spz1)

# shrinked results conditionwise
shres_spz1 <- lfcShrink(dds_spz1, type="apeglm", lfcThreshold=0,
                        coef="sample_type_test_vs_control")

shres_spz1_ashr <- lfcShrink(dds_spz1, type="ashr", lfcThreshold=0,
                            coef="sample_type_test_vs_control")

# padj < 0.05: subset results to return genes ----
# spz1 ashr
sig_shres_spz1_ashr <- shres_spz1_ashr %>%
  data.frame() %>%
  rownames_to_column(var="gene") %>% 
  as_tibble() %>% 
  filter(padj < 0.05)

# spz1 apeglm
sig_shres_spz1 <- shres_spz1 %>%
  data.frame() %>%
  rownames_to_column(var="gene") %>% 
  as_tibble() %>% 
  filter(padj < 0.05)
```

## In batch 2 of sporozoite stage
```{r batch2_spz}
# batch 2 spz ----
dds_spz2 <- DESeqDataSetFromMatrix(countData = countData_spz2, 
                                   colData = metaData_spz2,
                                   design = ~ sample_type, tidy = FALSE)

# filter out by at least 3 samples with a count of 10 or higher
keep_spz2 <- rowSums(counts(dds_spz2) >= 10) >= 3
dds_spz2 <- dds_spz2[keep_spz2,]

# apply DESEQ function
dds_spz2 <- DESeq(dds_spz2)

# shrinked results conditionwise
shres_spz2 <- lfcShrink(dds_spz2, type="apeglm", lfcThreshold=0,
                        coef="sample_type_test_vs_control")

shres_spz2_ashr <- lfcShrink(dds_spz2, type="ashr", lfcThreshold=0,
                             coef="sample_type_test_vs_control")

# padj < 0.05: subset results to return genes ----
# spz2 ashr
sig_shres_spz2_ashr <- shres_spz2_ashr %>%
  data.frame() %>%
  rownames_to_column(var="gene") %>% 
  as_tibble() %>% 
  filter(padj < 0.05)

# spz2 apeglm
sig_shres_spz2 <- shres_spz2 %>%
  data.frame() %>%
  rownames_to_column(var="gene") %>% 
  as_tibble() %>% 
  filter(padj < 0.05)
```

# Step 6: Analysis of Aging
```{r Aging}
# * DEA AGING ALL SAMPLES: ----
# build DEseq2 object for all the conditions
# make DESeq matrix
dds.aging <- DESeqDataSetFromMatrix(countData = countData, 
                                    colData = metaData,
                                    design = ~ infection_stage, tidy = FALSE)

# filter out by at least 3 samples with a count of 10 or higher ----
keep <- rowSums(counts(dds.aging) >= 10) >= 3
dds.aging <- dds.aging[keep,]

# apply DESEQ function ----
dds.aging <- DESeq(dds.aging)

# shrinked results conditionwise
shres_aging <- lfcShrink(dds.aging, type="apeglm", lfcThreshold=0,
                         coef=c("infection_stage_sporozoite_vs_oocyst"))

# old VS young
sig_shres_aging <- shres_aging %>%
  data.frame() %>%
  rownames_to_column(var="gene") %>% 
  as_tibble() %>% 
  filter(padj < 0.05)


# up
up_sig_shres_aging <- sig_shres_aging %>%
  data.frame() %>%
  as_tibble() %>% 
  filter(log2FoldChange > 0)


# down
down_sig_shres_aging <- sig_shres_aging %>%
  data.frame() %>%
  as_tibble() %>% 
  filter(log2FoldChange < 0)

```

# Step 7: Calculate DEGs among all sample replicates
```{r Dist_calc, results='hide'}
# * Plot distance among samples ----
# build DEseq2 object for all the conditions
# make DESeq matrix
dds.distance <- DESeqDataSetFromMatrix(countData = countData,
                                    colData = metaData,
                                    design = ~ 1, tidy = F)

# add phenofile cathegories interactions ("design" argument in the above function "DESeqDataSetFromMatrix")
dds.distance$group <- factor(paste0(dds.distance$sample_type, dds.distance$infection_stage, dds.distance$batch))
dds.distance$group <- relevel(dds.distance$group, ref = "controloocyst1")
design(dds.distance) <- ~ group
dds.distance$group <- gsub("3", "1", dds.distance$group)
dds.distance$group <- gsub("4", "2", dds.distance$group)
# Convert group to factor AFTER string substitution
dds.distance$group <- factor(dds.distance$group)

# filter out by at least 3 samples with a count of 10 or higher ----
keep <- rowSums(counts(dds.distance) >= 10) >= 3
dds.distance <- dds.distance[keep,]

# apply DESEQ function
dds.distance <- DESeq(dds.distance)

#unique(dds.distance$group) # see the final values
# Define your conditions
conditions <- rev(c("controloocyst", "controlsporozoite", "testoocyst", "testsporozoite"))
replicates <- 1:2  # assuming you have replicates 1 and 2

# Initialize a list to store all results
results_list <- list()

# Generate all unique non-self pairs
for (cond1 in conditions) {
  for (rep1 in replicates) {
    for (cond2 in conditions) {
      for (rep2 in replicates) {
# Skip self-comparisons
        if (!(cond1 == cond2 && rep1 == rep2)) {
# Create condition names
          name1 <- paste0(cond1, rep1)
          name2 <- paste0(cond2, rep2)
# Create a meaningful name for the result
          result_name <- paste0(gsub("control", "c", gsub("test", "", cond2)),
                                rep2,
                                gsub("control", "c", gsub("test", "", cond1)),
                                rep1)
# Perform the shrinkage
          cat("Processing contrast:", name2, "vs", name1, "...\n")
          res <- lfcShrink(dds.distance, type="ashr", lfcThreshold=0,
                           contrast=c("group", name2, name1))
# Store the result
          results_list[[result_name]] <- res
        }
      }
    }
  }
}


# Apply filtering and formatting to all results
sig_results <- lapply(results_list, function(res) {
  res %>%
    data.frame() %>%
    rownames_to_column(var = "gene") %>%
    as_tibble() %>%
    filter(padj < 0.05)
})


# Define all conditions (same order as your matrix)
conditions <- c("coocyst1", "coocyst2",
                "csporozoite1", "csporozoite2",
                "oocyst1", "oocyst2",
                "sporozoite1", "sporozoite2")

# initialize distance matrix
distance_matrix <- matrix(0, nrow=length(conditions), ncol=length(conditions),
                          dimnames=list(conditions, conditions))

# Fill the matrix in a loop
for (i in 1:length(conditions)) {
  for (j in 1:length(conditions)) {
    if (i == j) {
      distance_matrix[i,j] <- 0  # Diagonal is zero
    } else {
# Construct the result object name
      result_name <- paste0(conditions[j], conditions[i])

# Check if this comparison exists in sigresults
      if (exists(result_name, where = sig_results)) {
        res <- sig_results[[result_name]]
        distance_matrix[i,j] <- length(res$gene)  # Or your actual DEG count metric
      } else {
# Try reverse comparison if this one doesn't exist
        reverse_name <- paste0(conditions[i], conditions[j])
        if (exists(reverse_name, where = sig_results)) {
          res <- sig_results[[reverse_name]]
          distance_matrix[i,j] <- length(res$gene)  # Or your actual DEG count metric
        } else {
          distance_matrix[i,j] <- NA  # No comparison available
        }
      }
    }
  }
}


# Remove bottom part of the matrix (mirrored)
distance_matrix[lower.tri(distance_matrix)] <- 0

# save matrix
sep.reps.distance <- distance_matrix

# Melt the matrix
melted_matrix <- as.data.frame(as.table(as.matrix(sep.reps.distance))) # for a distance matrix with empty diagonal

# 5. Force column names to be exact
names(melted_matrix) <- c("Var1", "Var2", "Value")
```


# Step 8: Gene enrichment analysis of aging related genes
```{r Enrich}
# * aging immune DEGs family enrichment ----
# p-values are calculated using the upper-tail (right-tail) probability 
# P(X ≥ a). The FDR is adjusted across all families using the Benjamini-Hochberg method.

# * all DEGs
# Parameters
N <- 13845   # Total genes in Ag PEST genome
n <- 4397    # Number of DEGs in aging analysis
# Gene families: m = number in gene family, a = observed overlap
# families taken form Christophides 2002 (doi: 10.1126/science.1077136), some families mentioned by Christophides
# are absent because their gene name could not be found in the DEGs or in both degs and total gene list
# the counts have been done manually due to impossibility to always assign univocal strings to the gene names of a given family
families <- c("PGRP", "TEP", "GNBP", "SCR",
              "CTL", "GALE", "CLIP", # "FBN",
              "SRPN", "IAP", "TOLL + cactus", # "MyD88", "tube", "pelle",
              "REL", "Imd",
              "STAT", "PPO", "DEF", "CEC",
              "CASP")
m <- c(7, 14, 6, 18,
       21, 9, 57,
       14, 7, 8,
       2, 1,
       2, 9, 4, 3,
       12) # total number of genes in the family in Ag PEST genome (accessed 2025-05-02, GCA_000005575.2, AgamP4.14)

# NB: considered DEGs with logFC > |0.25|
a <- c(3, 4, 0, 8,
       8, 2, 29,
       8, 1, 1,
       0, 0,
       1, 4, 0, 0,
       0) # number of DEGs belonging to the family

a_up <- c(2, 3, 0, 3,
          4, 1, 27,
          8, 0, 0,
          0, 0,
          1, 3, 0, 0,
          0) # number of DEGs belonging to the family
a_down <- c(1, 1, 0, 5,
            4, 1, 2,
            0, 1, 1,
            0, 0,
            0, 1, 0, 0,
            0) # number of DEGs belonging to the family


# ALL GENES
# Calculate hypergeometric p-values
p_values <- phyper(a-1, m, N-m, n, lower.tail = FALSE)
# Calculate FDR (Benjamini-Hochberg)
fdr <- p.adjust(p_values, method = "fdr")
# Create a results table
results <- data.frame(
  Family = families,
  m = m,
  a = a,
  p_value = p_values,
  fdr = fdr
)
# Sort by FDR (ascending)
results <- results[order(results$fdr), ]

# * just upregulated genes
# Parameters
N <- 13845   # Total genes in Ag PEST genome
n <- 2166    # Number of urpeg DEGs in aging analysis
# ALL GENES
# Calculate hypergeometric p-values
p_values <- phyper(a_up-1, m, N-m, n, lower.tail = FALSE)
# Calculate FDR (Benjamini-Hochberg)
fdr <- p.adjust(p_values, method = "fdr")
# Create a results table
results <- data.frame(
  Family = families,
  m = m,
  a = a_up,
  p_value = p_values,
  fdr = fdr
)
# Sort by FDR (ascending)
results_up <- results[order(results$fdr), ]

# * just downregulated genes
# Parameters
N <- 13845   # Total genes in Ag PEST genome
n <- 2231    # Number of dowreg DEGs in aging analysis
# ALL GENES
# Calculate hypergeometric p-values
p_values <- phyper(a_down-1, m, N-m, n, lower.tail = FALSE)
# Calculate FDR (Benjamini-Hochberg)
fdr <- p.adjust(p_values, method = "fdr")
# Create a results table
results <- data.frame(
  Family = families,
  m = m,
  a = a_down,
  p_value = p_values,
  fdr = fdr
)
# Sort by FDR (ascending)
results_down <- results[order(results$fdr), ]
# Print the results

```

# Step 9: Plots {.tabset}
```{r Plots_load}
all_aging_info <- read.csv(file.path(projDir,"/Data/Bulk/FINAL_aging_genes.csv"), header = T, row.names = 1)
```

## PCA {.tabset}
```{r Plots PCA}
rld <- rlog(dds, blind = FALSE)

# example pca
# PC1,3
pcaData_age <- alternative.pca1_3(rld, intgroup= "infection_stage", returnData=TRUE)
percentVar_age <- round(100 * attr(pcaData_age, "percentVar"))
ggplot(pcaData_age, aes(PC1, PC2, col = infection_stage, shape = infection_stage)) +
  geom_point(size=2, shape=16) +
  theme(
    axis.text = element_text(size = 10),  # Smaller axis labels
    axis.title = element_text(size = 10),  # Smaller PC_1 and PC_2 labels
    legend.title = element_text(size = 10),  # Smaller legend title
    legend.text = element_text(size = 10),  # Smaller legend text
    panel.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    plot.background = element_rect(fill = "white", color = NA),
    axis.line = element_line(colour = "black")
  ) +
  xlab(paste0("PC1: ",percentVar_age[1],"% variance")) +
  ylab(paste0("PC3: ",percentVar_age[2],"% variance")) +
  labs(colour="age") +
  xlim(-30, 30) +
  ylim(-30, 30) +
  scale_color_manual(labels = c("14 days old", "22 days old"), values = c("orange2", "royalblue")) +
  coord_fixed()
```

## Heatmap of top differentially expressed genes (adjusted p-value < 0.05) between 14. and 22 day old females.
```{r Plots_heat, fig.width=10, fig.height=5}
# top FC and FDR genes to plot ----
# Add a column to the data frame to specify if they are UP- or DOWN- regulated (log2fc respectively positive or negative)
all_aging_info$diffexpressed <- "NO"
# if log2Foldchange > 0.5 and pvalue < 0.05, set as "UP"
all_aging_info$diffexpressed[all_aging_info$log2FoldChange > 1.5 & all_aging_info$padj < 0.05] <- "UP"
# if log2Foldchange < -0.5 and pvalue < 0.05, set as "DOWN"
all_aging_info$diffexpressed[all_aging_info$log2FoldChange < -1.5 & all_aging_info$padj < 0.05] <- "DOWN"

# which genes are there
diffexpressed_top_genes <- all_aging_info %>% 
  filter(diffexpressed != "NO")

# get genes to plot
# Filter the rows in diffexpressed_top_genes where p-value is < 10^20
genes_to_plot_vp <- diffexpressed_top_genes[diffexpressed_top_genes$padj < 10^(-19), ]
# * check if they are all present in all_aging_info ----
plot_aging_strength <- Reduce(intersect, list(all_aging_info$gene, genes_to_plot_vp$gene))


# heatmap top DEGs ----
# plot trend of chosen genes of interest in the downloaded DEGs of aging
# Define the list of genes
gene_list <- rownames(genes_to_plot_vp)
# in this case I put DEGs interesting in aging

# extract the assay matrix from the transformed object
mat.age <- assay(rlog(dds.aging, blind = FALSE))
# Compute z-scores for each gene (row-wise scaling)
z_scores.age <- t(scale(t(mat.age)))

# extract data matrix
expr_data <- z_scores.age

# Filter for genes and columns of interest
checked_data <- expr_data[rownames(expr_data) %in% gene_list, grep("^oocy|^sporo", colnames(expr_data))]

# Check if checked_data has valid dimensions
if (nrow(checked_data) == 0 | ncol(checked_data) == 0) {
  stop("No matching genes or samples found.")
}

# Reorder the data matrix to match the order in gene_list
ordered_data <- checked_data[match(gene_list, rownames(checked_data)), ]

# get groups of virgin and mated samples
oo_data <- checked_data[, grep("^oocy", colnames(checked_data))]
spz_data <- checked_data[, grep("^sporo", colnames(checked_data))]

# to order samples based on their average expression ofo the different genes
# in the two groups, perform hierarchical clustering on each group separately
oo_clustering <- hclust(dist(t(oo_data)))
spz_clustering <- hclust(dist(t(spz_data)))

# combine in the clustering order
ordered_cols <- c(
  colnames(oo_data)[oo_clustering$order],
  colnames(spz_data)[spz_clustering$order]
)
# reorder the data matrix
ordered_data <- checked_data[, ordered_cols]
# reorder rows based on their log2FC (i.e. their order in gene_list)
ordered_data <- ordered_data[match(gene_list, rownames(ordered_data)), ]

# transform data for ggplot2
long_data <- as.data.frame(ordered_data) %>%
  rownames_to_column(var = "Gene") %>%
  melt(id.vars = "Gene", variable.name = "Sample", value.name = "Expression")
# Create a dataframe for sample type
sample_type_df <- data.frame(
  Sample = colnames(dds.aging),
  sample_type = dds.aging$sample_type
)
# Create a dataframe for sample type
sample_type_df <- data.frame(
  Sample = colnames(dds.aging),
  sample_type = dds.aging$sample_type,
  Group = ifelse(grepl("^oocy", colnames(dds.aging)), "14 d.o.", "22 d.o.")
)

# Add sample_type to long_data
long_data <- merge(long_data, sample_type_df, by = "Sample")
# add sample group (young/old) for annotation
long_data <- long_data %>%
  mutate(Group = ifelse(grepl("^oocy", Sample), "14 d.o.", "22 d.o."))


# plot heatmap
# ggplot with gene names and sample bar
ggplot() +
  # Sample type color bar
  geom_tile(data = distinct(sample_type_df),
            aes(x = factor(Sample, levels = ordered_cols), 
                y = "Sample Type", 
                fill = sample_type)) +
  scale_fill_manual(name = "Sample Type",
                    values = c("control" = "magenta4", "test" = "#DC3220")) +
  ggnewscale::new_scale_fill() +
  
  # Expression heatmap
  geom_tile(data = long_data,
            aes(x = factor(Sample, levels = ordered_cols), 
                y = factor(Gene, levels = rev(gene_list)), 
                fill = Expression)) +
  scale_fill_viridis(option = "B", name = "Expression") +
  
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
    axis.text.y = element_text(size = 9),
    axis.title = element_blank(),
    legend.position = "right"
  ) +
  facet_grid(~Group, scales = "free_x", space = "free_x") +
  labs(title = "Gene Expression Heatmap (14 d.o. and 22 d.o. Groups)") +
  scale_y_discrete(labels = c("Sample Type" = "", gene_list))  # Hide label on the annotation bar

```

## Number of DEGs among all sample replicates  {.tabset}
```{r Plots_distance}
# 6. Plot with guaranteed correct column names
ggplot(melted_matrix, aes(x=Var1, y=Var2, fill=Value)) +
  geom_tile(color="white") +
  scale_fill_viridis_c(name="# DEGs", option="plasma", na.value="white") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle=45, hjust=1, size=10),
    axis.text.y = element_text(size=10)
  ) +
  labs(x="", y="") +
  coord_fixed()

```

